---
title: "GPU-accelerated SEACells"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{GPU-accelerated SEACells}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Gregor Lueg"
---

## Intro

SEACells ([Persad et al., 2023](https://www.nature.com/articles/s41587-023-01716-9))
finds meta cells by kernel archetypal analysis. The expensive part is the inner
Frank-Wolfe loop that updates the archetype matrix `B`: every iteration scans a
gradient over all cells for every archetype and takes an argmin. That scan is
where the runtime goes, and it is what `generate_seacells_gpu_sc()` moves onto
the GPU.

Everything else stays on the CPU: the kNN graph, the kernel matrix, the
archetype initialisation, the `A` update and the aggregation into pseudo-bulk
counts. So the speedup is bounded by how much of the run the B-argmin owns. That
share grows with `n_sea_cells`, because the gradient scan is `O(n_cells x k)`
while most of the CPU-side work is not. Ask for 50 meta cells and the win is
small. Ask for a few thousand and it gets interesting.

If you have not seen the CPU version, read the
[bixverse meta cell vignette](https://gregorlueg.github.io/bixverse/articles/single_cell_meta_cells.html)
first. The params, the result class and everything downstream are identical
here.

```{r}
#| label: setup
library(bixverse)
library(bixverse.gpu)
library(data.table)
library(magrittr)
```

::: {.callout-note}
Vignettes were built locally on a MacBook Pro M1 Max. The GH runners were just
too slow and do not have proper GPU support. This gives an idea of speed on a
decent, but older machine.
:::

## Rebuilding a processed PBMC3k object

Same object as the other GPU vignettes. The chunk is folded because we care
about meta cells, not QC.

```{r}
#| label: rebuild-object
#| code-fold: true
#| code-summary: "Rebuild the PBMC3k object (click to expand)"
pbmc3k_path <- download_pbmc3k()
tempdir_pbmc <- file.path(tempdir(), "gpu_metacells")
dir.create(tempdir_pbmc, showWarnings = FALSE, recursive = TRUE)

sc_object <- SingleCells(dir_data = tempdir_pbmc)
mtx_io_params <- get_cell_ranger_params(pbmc3k_path)

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = mtx_io_params,
  mtx_streaming = FALSE,
  .verbose = FALSE
)

setnames_sc(sc_object, table = "var", old = "column1", new = "gene_symbol")

var <- get_sc_var(sc_object)
ensembl_to_symbol <- setNames(var$gene_symbol, var$gene_id)

gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_symbol), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_symbol), gene_id]
)
sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  streaming = FALSE,
  .verbose = FALSE
)

qc_df <- sc_object[[c("cell_id", "lib_size", "nnz", "MT")]]
metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)
directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)
qc <- run_cell_qc(
  metrics = metrics,
  cells_to_keep = get_cells_to_keep(sc_object),
  directions = directions,
  threshold = 3
)
sc_object[["outlier"]] <- qc$combined
cells_to_keep <- qc_df[!qc$combined, cell_id]
sc_object <- set_cells_to_keep(sc_object, cells_to_keep)

sc_object <- find_hvg_sc(sc_object, hvg_no = 2000L, .verbose = FALSE)
sc_object <- calculate_pca_sc(sc_object, no_pcs = 30L, .verbose = FALSE)
sc_object <- find_neighbours_sc(
  sc_object,
  neighbours_params = params_sc_neighbours(knn = list(k = 15L)),
  .verbose = FALSE
)
```

## Running it

One call, and the params are the same `params_sc_seacells()` the CPU version
takes. PBMC3k is small, so aim for one meta cell per ~15 cells.

```{r}
#| label: gpu-seacells
seacell_params <- params_sc_seacells(
  n_sea_cells = 150L,
  knn = list(k = 25L)
)

gpu_time <- system.time({
  mc_gpu <- generate_seacells_gpu_sc(
    sc_object,
    seacell_params = seacell_params,
    .verbose = TRUE
  )
})

mc_gpu
```

The result is a `MetaCells` object, byte for byte the same shape as the CPU one.
Assignments live in `@original_assignment`, the RSS trace and the archetype cell
indices in `@other_data`.

```{r}
#| label: gpu-seacells-inspect
head(mc_gpu[[]], 3L)

tail(mc_gpu@other_data$rss, 5L)
```

### Knobs that matter on large data

Two params change the picture once you leave toy datasets:

- **`n_landmarks`** switches archetype initialisation to a Nystroem
  approximation on a landmark subsample. Without it, initialisation is
  quadratic in cells and starts to dominate the run well before the
  Frank-Wolfe loop does.
- **`pruning`** drops tiny values during the Frank-Wolfe updates. It costs a bit
  of numerical fidelity and buys a lot of memory, which is what makes million-
  cell runs feasible at all. `pruning_threshold` controls how aggressive it is.

```{r}
#| label: gpu-seacells-knobs
#| eval: false
seacell_params_large <- params_sc_seacells(
  n_sea_cells = 5000L,
  n_landmarks = 20000L,
  pruning = TRUE,
  pruning_threshold = 1e-7,
  knn = list(k = 25L)
)
```

## CPU versus GPU

Same object, same params, same seed.

```{r}
#| label: cpu-seacells
cpu_time <- system.time({
  mc_cpu <- generate_seacells_sc(
    sc_object,
    seacell_params = seacell_params,
    .verbose = FALSE
  )
})

data.table(
  version = c("CPU", "GPU"),
  seconds = round(c(cpu_time[["elapsed"]], gpu_time[["elapsed"]]), 2)
)[, speed_up := round(seconds[1] / seconds, 2)][]
```

The two do not produce bit-identical assignments. GPU reduction ordering breaks
ties differently, so a cell sitting almost equidistant between two archetypes
can land either way. What should agree is the structure. Compare how the meta
cells partition the cells:

```{r}
#| label: cpu-vs-gpu
mc_sizes <- data.table(
  version = c("CPU", "GPU"),
  n_meta_cells = c(nrow(mc_cpu[[]]), nrow(mc_gpu[[]])),
  median_cells = c(
    median(mc_cpu[[]]$no_originating_cells),
    median(mc_gpu[[]]$no_originating_cells)
  ),
  final_rss = round(c(
    tail(mc_cpu@other_data$rss, 1L),
    tail(mc_gpu@other_data$rss, 1L)
  ), 2)
)

mc_sizes
```

On a dataset this size the GPU rarely blows the CPU out of the water. PBMC3k is
around 2.7k cells and 150 archetypes, which is not enough work to hide the
launch overhead, and the CPU version parallelises the same scan across all
performance cores. The GPU path earns its keep further up the scale, where
`n_cells x k` gets large enough that the argmin genuinely dominates.

## Feeding into SCENIC

The resulting `MetaCells` goes straight into the GPU SCENIC driver, no
conversion step.

```{r}
#| label: metacells-to-scenic
#| eval: false
scenic_res <- scenic_grn_sc_gpu(
  object = mc_gpu,
  tf_ids = tf_dt$gene_id,
  scenic_params = params_scenic(learner_type = "extratrees"),
  .verbose = TRUE
)
```

See the [GPU SCENIC vignette](gpu_scenic.html) for the full workflow.

## What is next

The Frank-Wolfe `A` update and the RSS evaluation are still CPU-side, and both
are matrix products against the same kernel matrix that already lives near the
device. Moving them would cut the CPU/GPU round trips per outer iteration.
Watch this space.
