---
title: "GPU-accelerated NEBULA"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{GPU-accelerated NEBULA}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Gregor Lueg"
---

## Intro

NEBULA fits a negative binomial gamma mixed model per gene, with the donor as
a random effect. Why you'd want that over a Wilcoxon test is covered in the
[differential expression vignette](https://gregorlueg.github.io/bixverse/articles/differential_expression.html)
in `bixverse`. Read that first. This one only asks two questions: does the GPU
version give you the same answer, and is it faster?

`nebula_gpu_sc()` is a drop-in for `bixverse::nebula_sc()`. Same arguments,
same `ScNebula` result. Stage two of NEBULA, the per-gene penalised fits, runs
on the device via [cubecl](https://github.com/tracel-ai/cubecl). Cell ordering,
gene batching, the dispersion shrinkage and the Wald test are the CPU code.

Two differences are worth knowing about:

- The device fits run in `f32` and the host finishes each one in `f64`. The
  numbers land close to the CPU ones, not on them.
- REML is not implemented on the device. `params_nebula_gpu()` is
  `bixverse::params_nebula()` without the `reml` knob.

```{r}
#| label: setup
#| message: false
#| warning: false
library(bixverse)
library(bixverse.gpu)
library(data.table)
library(ggplot2)
library(SingleCellExperiment)
```

## The data

Same set-up as the CPU vignette: the [Kang, et al.](https://doi.org/10.1038/nbt.4042)
PBMCs, eight lupus donors, each split into an IFN-β stimulated and a control
arm. Doublets out, then CD14+ monocytes and their 500 most variable genes.

```{r}
#| label: load-data
#| code-fold: true
#| code-summary: "Load and subset the Kang data (click to expand)"
sce <- qs2::qs_read(download_kang_pbmc())
sce <- sce[, sce$multiplets == "singlet" & !is.na(sce$cell)]

dir_kang <- file.path(tempdir(), "kang_gpu")
dir.create(dir_kang, showWarnings = FALSE, recursive = TRUE)

sc_object <- SingleCells(dir_data = dir_kang)
sc_object <- load_sce(
  object = sc_object,
  sce = sce,
  sc_qc_param = params_sc_min_quality(
    min_unique_genes = 200L,
    min_lib_size = 500L,
    min_cells = 20L,
    target_size = 1e4
  ),
  .verbose = FALSE
)

mono <- SingleCellsSubset(
  sc_object,
  grouping_column = "cell",
  group = "CD14+ Monocytes"
)

mono <- find_hvg_sc(mono, hvg_no = 500L, .verbose = FALSE)
hvgs <- get_gene_names_from_idx(mono, get_hvg(mono), rust_based = TRUE)
```

```{r}
#| label: design
mono
```

## The real contrast

Stimulated against control, donor as the random effect. Run both, time both.

```{r}
#| label: real-cpu
time_cpu <- system.time(
  res_cpu <- nebula_sc(
    object = mono,
    subject_col = "ind",
    design = ~stim,
    genes_to_use = hvgs,
    .verbose = TRUE
  )
)
```

```{r}
#| label: real-gpu
time_gpu <- system.time(
  res_gpu <- nebula_gpu_sc(
    object = mono,
    subject_col = "ind",
    design = ~stim,
    genes_to_use = hvgs,
    .verbose = TRUE
  )
)
```

```{r}
#| label: real-timing
data.table(
  backend = c("CPU", "GPU"),
  seconds = c(time_cpu[["elapsed"]], time_gpu[["elapsed"]])
)
```

### How close are they?

Join the two result tables on the gene and look at the worst absolute gap per
column. Relative gaps are no use here: a gene with an effect of `1e-3` turns an
absolute gap in the fourth decimal into a 70% relative one.

```{r}
#| label: real-diff
res_both <- merge(
  res_cpu$results,
  res_gpu$results,
  by = "gene_id",
  suffixes = c("_cpu", "_gpu")
)

cols <- c(
  "log_fc",
  "z",
  "p_value",
  "fdr",
  "subject_overdispersion",
  "cell_overdispersion"
)

rbindlist(lapply(cols, \(col) {
  a <- res_both[[paste0(col, "_cpu")]]
  b <- res_both[[paste0(col, "_gpu")]]
  data.table(
    column = col,
    max_abs_diff = max(abs(a - b)),
    pearson = cor(a, b)
  )
}))
```

```{r}
#| label: real-scatter
#| fig-width: 8
#| fig-height: 4
plot_dt <- rbind(
  res_both[, .(stat = "log_fc", cpu = log_fc_cpu, gpu = log_fc_gpu)],
  res_both[, .(stat = "z", cpu = z_cpu, gpu = z_gpu)]
)

ggplot(plot_dt, aes(x = cpu, y = gpu)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = 2) +
  geom_point(size = 0.8, alpha = 0.6) +
  facet_wrap(~stat, scales = "free") +
  labs(x = "CPU", y = "GPU") +
  theme_minimal()
```

What matters in the end is the calls. Same genes at FDR <= 0.05?

```{r}
#| label: real-calls
res_both[, .(
  genes = .N,
  sig_cpu = sum(fdr_cpu <= 0.05),
  sig_gpu = sum(fdr_gpu <= 0.05),
  sig_both = sum(fdr_cpu <= 0.05 & fdr_gpu <= 0.05),
  bound_agree = sum(sigma_at_bound_cpu == sigma_at_bound_gpu),
  convergence_agree = sum(convergence_cpu == convergence_gpu)
)]
```

Every CPU call is a GPU call and vice versa. The top of the list, ranked by the
CPU's `|z|`, side by side:

```{r}
#| label: real-top
head(
  res_both[
    order(-abs(z_cpu)),
    .(gene_id, log_fc_cpu, log_fc_gpu, z_cpu, z_gpu)
  ],
  8
)
```

The interferon response, ISG15, APOBEC3A and CXCL10, identical to the printed
precision. The one place the two disagree is the convergence code, on a handful
of genes:

```{r}
#| label: real-convergence
res_both[
  convergence_cpu != convergence_gpu,
  .(gene_id, convergence_cpu, convergence_gpu, z_cpu, z_gpu, fdr_cpu, fdr_gpu)
]
```

The codes swap between `1` and `-10` in both directions. Both sit above the
`-20` line that marks a likely failure, and `z` and the FDR on these genes
still agree to five or six significant digits. Nothing here changes a call.

## No contrast at all

The CPU vignette's key check: control cells only, eight donors split four
against four at random. Any call is a false positive. If the GPU's `f32` fits
nudged the overdispersions the wrong way, this is where extra calls would show
up.

```{r}
#| label: null-setup
obs_parent <- sc_object[[]]
donors <- sort(unique(obs_parent$ind))
group_a <- donors[c(1, 3, 5, 7)]

sc_object[["cell_stim"]] <- paste(obs_parent$cell, obs_parent$stim, sep = "__")
sc_object[["fake"]] <- ifelse(obs_parent$ind %in% group_a, "a", "b")

mono_ctrl <- SingleCellsSubset(
  sc_object,
  grouping_column = "cell_stim",
  group = "CD14+ Monocytes__ctrl"
)
```

```{r}
#| label: null-run
null_cpu <- nebula_sc(
  object = mono_ctrl,
  subject_col = "ind",
  design = ~fake,
  genes_to_use = hvgs,
  .verbose = TRUE
)

null_gpu <- nebula_gpu_sc(
  object = mono_ctrl,
  subject_col = "ind",
  design = ~fake,
  genes_to_use = hvgs,
  .verbose = TRUE
)

data.table(
  backend = c("CPU", "GPU"),
  false_positives = c(
    null_cpu$results[fdr <= 0.05, .N],
    null_gpu$results[fdr <= 0.05, .N]
  )
)
```

## Timing

On this data, `r length(hvgs)` genes over `r res_cpu$params$n_cells` cells,
the CPU took `r round(time_cpu[["elapsed"]], 1)` s and the GPU
`r round(time_gpu[["elapsed"]], 1)` s. One run each on an Apple Silicon
laptop, device set-up included. Across renders of this vignette the order has
flipped both ways, so call it a wash at this size. The whole sweep is seconds
either way, and stage one, the batching and the Wald test stay on the CPU in
both, so the GPU only ever takes over part of the run. Larger cell counts and
longer gene lists are where stage two grows, but that isn't measured here.
Benchmark on your own data before committing to the GPU path.

## Where next

REML on the device isn't there yet; use the CPU path if you need it. NEBULA on
`MetaCells` stays CPU-only for now too.
