---
title: "GPU-accelerated SCENIC"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{GPU-accelerated SCENIC}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Gregor Lueg"
---

## Intro

This vignette walks through the GPU-accelerated version of the SCENIC gene
regulatory network inference in `bixverse`. If you have not read the
[CPU SCENIC vignette](https://gregorlueg.github.io/bixverse/articles/bag_of_genes_single_cells.html)
first, do so. That vignette explains the biology and the CPU code path.
This one focuses on what the GPU version buys you and how the API differs.

Behind `scenic_grn_sc_gpu()` sits a wave-scheduled, multi-tree GPU driver
that dispatches ExtraTrees and RandomForest fits to the wgpu backend via
[cubecl](https://github.com/tracel-ai/cubecl). GRNBoost2 stays on the CPU
because gradient boosting doesn't map cleanly onto the same multi-output
kernel design.

The `bixverse.gpu` package provides:

- **`scenic_grn_sc_gpu()`** for `SingleCells` and `MetaCells` objects.
  Streaming and non-streaming drivers on the `SingleCells` path; a single
  in-memory driver on the `MetaCells` path.
- A single GPU-specific knob, `wave_byte_budget`, that caps VRAM use of
  the per-wave histogram tensors. Default is 4 GiB. Shrink on 8 GB
  adapters shared with other workloads; raise on 16 GB+ adapters to keep
  the wave at 8.

Everything else, the params, the target-gene shortlisting, the downstream
motif enrichment, is imported straight from `bixverse` — the GPU version
reuses the same `params_scenic()`, the same `scenic_gene_filter_sc()`, the
same `ScenicGrn` result class, so downstream code (`identify_tf_to_genes`,
`tf_to_genes_correlations`, `tf_to_genes_motif_enrichment`) is identical.

```{r}
#| label: setup
library(bixverse)
library(bixverse.gpu)
library(data.table)
library(magrittr)
```

::: {.callout-note}
Vignettes are built on GitHub CI/CD runners with no GPU. GPU code falls
back to software Vulkan (lavapipe), so `N` is dialled down here for build
time. On real hardware SCENIC eats much bigger data sets without breaking
a sweat.
:::

## Rebuilding a processed PBMC3k object

We reconstruct the same `SingleCells` object used in the CPU SCENIC
vignette. The chunk is folded because we care about SCENIC, not QC.

```{r}
#| label: rebuild-object
#| code-fold: true
#| code-summary: "Rebuild the PBMC3k object (click to expand)"
pbmc3k_path <- download_pbmc3k()
tempdir_pbmc <- file.path(tempdir(), "gpu_scenic")
dir.create(tempdir_pbmc, showWarnings = FALSE, recursive = TRUE)

sc_object <- SingleCells(dir_data = tempdir_pbmc)
mtx_io_params <- get_cell_ranger_params(pbmc3k_path)

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = mtx_io_params,
  mtx_streaming = FALSE,
  .verbose = FALSE
)

setnames_sc(sc_object, table = "var", old = "column1", new = "gene_symbol")

var <- get_sc_var(sc_object)
ensembl_to_symbol <- setNames(var$gene_symbol, var$gene_id)
symbol_to_ensembl <- setNames(var$gene_id, var$gene_symbol)

gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_symbol), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_symbol), gene_id]
)
sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  streaming = FALSE,
  .verbose = FALSE
)

qc_df <- sc_object[[c("cell_id", "lib_size", "nnz", "MT")]]
metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)
directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)
qc <- run_cell_qc(
  metrics = metrics,
  cells_to_keep = get_cells_to_keep(sc_object),
  directions = directions,
  threshold = 3
)
sc_object[["outlier"]] <- qc$combined
cells_to_keep <- qc_df[!qc$combined, cell_id]
sc_object <- set_cells_to_keep(sc_object, cells_to_keep)

sc_object <- find_hvg_sc(sc_object, hvg_no = 2000L, .verbose = FALSE)
sc_object <- calculate_pca_sc(sc_object, no_pcs = 30L, .verbose = FALSE)
sc_object <- find_neighbours_sc(
  sc_object,
  neighbours_params = params_sc_neighbours(knn = list(k = 15L)),
  .verbose = FALSE
)
```

## Gene filtering

Same as the CPU workflow. The gene filter is a cheap min-counts /
min-cells scan; there's no GPU version because there's no need for one.

```{r}
#| label: scenic-gene-filter
scenic_genes <- scenic_gene_filter_sc(
  object = sc_object,
  scenic_params = params_scenic(min_counts = 50L),
  .verbose = FALSE
)

length(scenic_genes)
```

## Transcription factor list

Aerts lab curated list, mapped to Ensembl.

```{r}
#| label: scenic-tfs
tf_dt <- data.table::fread(
  "https://resources.aertslab.org/cistarget/tf_lists/allTFs_hg38.txt",
  header = FALSE,
  col.names = "tf"
)
tf_dt[, gene_id := symbol_to_ensembl[tf]]
tf_dt <- tf_dt[!is.na(gene_id)]

nrow(tf_dt)
```

## GRN inference on the GPU

Same signature as `scenic_grn_sc()` from `bixverse`, with the extra
`wave_byte_budget` knob. Everything else, `scenic_params`, `tf_ids`,
`genes_to_take`, works the same way.

### ExtraTrees

`extratrees` is the recommended default on GPU. Random split thresholds
mean each level dispatches a single split-scoring kernel, and the driver
saturates the device with wide waves.

```{r}
#| label: gpu-scenic-et
scenic_res_et <- scenic_grn_sc_gpu(
  object = sc_object,
  tf_ids = tf_dt$gene_id,
  genes_to_take = scenic_genes,
  scenic_params = params_scenic(
    learner_type = "extratrees",
    gene_batch_size = 64L
  ),
  wave_byte_budget = 4 * 1024^3,
  .verbose = TRUE
)

scenic_res_et
```

### RandomForest

`randomforest` uses exhaustive split evaluation. It's a bit slower per
tree than ExtraTrees but often lands on tighter importance rankings.

```{r}
#| label: gpu-scenic-rf
scenic_res_rf <- scenic_grn_sc_gpu(
  object = sc_object,
  tf_ids = tf_dt$gene_id,
  genes_to_take = scenic_genes,
  scenic_params = params_scenic(
    learner_type = "randomforest",
    gene_batch_size = 64L
  ),
  .verbose = TRUE
)

scenic_res_rf
```

### GRNBoost2

Not available on GPU. The wrapper fails fast with a clear message before
touching the device.

```{r}
#| label: gpu-scenic-grnboost2
#| error: true
scenic_grn_sc_gpu(
  object = sc_object,
  tf_ids = tf_dt$gene_id,
  genes_to_take = scenic_genes,
  scenic_params = params_scenic(learner_type = "grnboost2"),
  .verbose = FALSE
)
```

Fall back to `bixverse::scenic_grn_sc()` if you need it.

### Streaming vs non-streaming

On disk-backed `SingleCells` objects both drivers exist. Non-streaming
loads the full sparse target column vector into host RAM up front and
dispatches wave by wave. Streaming reads targets in I/O chunks and
dispatches per chunk, capping peak host memory at roughly one chunk of
sparse columns.

Set `streaming = FALSE` to force the non-streaming path when host RAM
isn't the limiting factor:

```{r}
#| label: gpu-scenic-streaming
#| eval: false
scenic_res_streaming <- scenic_grn_sc_gpu(
  object = sc_object,
  tf_ids = tf_dt$gene_id,
  genes_to_take = scenic_genes,
  scenic_params = params_scenic(
    learner_type = "extratrees",
    gene_batch_size = 64L
  ),
  streaming = TRUE,
  .verbose = TRUE
)
```

Leave `streaming = NULL` to let the same `auto_streaming()` heuristic that
the CPU version uses pick for you.

## TF-to-gene refinement

Because the result is a `ScenicGrn` object, the CPU downstream steps drop
straight in.

```{r}
#| label: gpu-scenic-tf-to-gene
scenic_res_et <- identify_tf_to_genes(
  scenic_res_et,
  n_sd = 1,
  .verbose = TRUE
)

scenic_res_et <- tf_to_genes_correlations(
  x = scenic_res_et,
  object = sc_object,
  cor_filter = 0.01,
  .verbose = TRUE
)

tf_gene_dt <- get_tf_to_gene(scenic_res_et)
tf_gene_dt[, tf_symbol := ensembl_to_symbol[tf]]
tf_gene_dt[, gene_symbol := ensembl_to_symbol[gene]]

head(tf_gene_dt[order(-importance)], 5L)
```

## MetaCells path

The GPU driver also dispatches on `MetaCells`. Counts already live in
memory as a sparse CSC block, so the streaming flag has no effect (and is
silently ignored). Build the metacells the same way as the CPU vignette:

```{r}
#| label: gpu-scenic-metacells
mc_object <- generate_bt_meta_cells_sc(
  sc_object,
  sc_meta_cell_params = params_sc_bt_metacells(
    target_no_metacells = 200L
  ),
  .verbose = FALSE
)

scenic_res_mc <- scenic_grn_sc_gpu(
  object = mc_object,
  tf_ids = tf_dt$gene_id,
  scenic_params = params_scenic(
    learner_type = "extratrees",
    min_counts = 50L,
    min_cells = 0.02,
    gene_batch_size = 64L
  ),
  .verbose = TRUE
)

scenic_res_mc
```

Same result class, same downstream methods. `identify_tf_to_genes()`,
`tf_to_genes_correlations()` and `tf_to_genes_motif_enrichment()` all work
unchanged.

## CisTarget motif enrichment

Nothing changes on the CisTarget side. Same rankings, same annotations,
same result. Not evaluated by default because the reference files are
around 1.5 GB.

```{r}
#| label: gpu-scenic-cistarget
#| eval: false
paths <- download_cistarget_hg38()

rankings <- read_motif_ranking(paths$rankings)
annotations <- read_motif_annotation_file(paths$motif_annotations)

scenic_res_et <- tf_to_genes_motif_enrichment(
  x = scenic_res_et,
  motif_rankings = rankings,
  annot_data = annotations,
  gene_id_to_symbol = ensembl_to_symbol,
  cis_target_params = params_cistarget(nes_threshold = 3.0),
  only_high_conf_tf = FALSE,
  .verbose = TRUE
)

cistarget_dt <- get_cistarget_res(scenic_res_et)
head(cistarget_dt[order(-nes)])
```

## Choosing between CPU and GPU

- Small data (a few thousand cells): the CPU path is fine and the GPU
  setup cost eats most of the win.
- Large data (tens of thousands of cells, thousands of TFs, hundreds of
  target batches): GPU pays off. ExtraTrees on GPU is where the biggest
  wins live because random-split evaluation maps cleanly onto the
  segmented histogram kernels.
- On Apple Silicon the GPU rarely blows the CPU out of the water because
  the CPU already has fast unified-memory access. On discrete GPUs with
  16 GB+ VRAM the gap widens quickly.
- If you need GRNBoost2, stay on `bixverse::scenic_grn_sc()`. GRNBoost2's
  single-target histogram-subtraction algorithm doesn't share structure
  across targets and gains nothing from a GPU port.

That's the whole GPU SCENIC surface. If you land here from the CPU
vignette, everything you know about `scenic_params`, `tf_ids` and the
downstream `ScenicGrn` methods keeps working.
