---
title: "GPU-accelerated Scrublet"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{GPU-accelerated Scrublet}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Gregor Lueg"
---

## Intro

Scrublet ([Wolock et al., 2019](https://www.sciencedirect.com/science/article/pii/S2405471218304745))
detects doublets by simulating them. It combines random pairs of observed cells
into artificial doublets, projects observed and simulated cells into a shared
PC space, builds a kNN graph over the lot, and scores each observed cell by how
many of its neighbours are simulated doublets. Otsu's method then cuts the
score distribution into singlets and doublets.

Three stages move to the GPU in `scrublet_gpu_sc()`: the randomised sparse SVD
of the observed cells, the projection of the simulated doublets into that PC
space, and the kNN over the combined embedding. HVG selection, the doublet
simulation itself, the kNN classifier and the Otsu threshold stay on the CPU.

So the win is bounded by how much of the run those three own, and that share
grows with cell count. The combined embedding is
`(1 + sim_doublet_ratio) * n_cells` rows tall, and an exhaustive kNN over it is
quadratic. At 5k cells the setup cost dominates and you will not see much. At
100k it gets painful on the CPU and the GPU starts to matter.

If you have not seen the CPU version, read the
[bixverse doublet detection vignette](https://gregorlueg.github.io/bixverse/articles/doublet_detection.html)
first. The result class and everything downstream of it are identical here.

```{r}
#| label: setup
library(bixverse)
library(bixverse.gpu)
library(data.table)
library(magrittr)
```

::: {.callout-note}
Vignettes were built locally on a MacBook Pro M1 Max. The GH runners were just
too slow and do not have proper GPU support. This gives an idea of speed on a
decent, but older machine.
:::

## The data

PBMCs with demuxlet ground truth calls. Each barcode is classified as a singlet
(`SNG`) or a doublet (`DBL`), so there is something real to score against.

```{r}
#| label: load-data
#| code-fold: true
#| code-summary: "Load the demuxlet PBMCs (click to expand)"
doublet_path <- download_demuxlet_pbmc()

tempdir_doublet <- file.path(tempdir(), "demuxlet_bixverse_gpu")
dir.create(tempdir_doublet, showWarnings = FALSE, recursive = TRUE)

demuxlet_data <- fread(file.path(doublet_path, "demuxlet_calls.tsv"))

sc_object <- SingleCells(dir_data = tempdir_doublet)

mtx_io_params <- get_cell_ranger_params(doublet_path)
mtx_io_params$cells_as_rows <- TRUE

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = mtx_io_params,
  streaming = 0L
)

sc_object
```

A small helper for precision, recall and F1 against the demuxlet calls.

```{r}
#| label: metrics-helper
doublet_metrics <- function(
  predicted,
  actual,
  pos_predicted = TRUE,
  pos_actual = "DBL"
) {
  tp <- sum(predicted == pos_predicted & actual == pos_actual)
  fp <- sum(predicted == pos_predicted & actual != pos_actual)
  fn <- sum(predicted != pos_predicted & actual == pos_actual)
  precision <- tp / (tp + fp)
  recall <- tp / (tp + fn)
  f1 <- 2 * (precision * recall) / (precision + recall)
  list(precision = precision, recall = recall, f1 = f1)
}
```

## Running it

```{r}
#| label: gpu-scrublet
gpu_params <- params_scrublet_gpu(expected_doublet_rate = 0.12)

gpu_time <- system.time({
  scrublet_gpu <- scrublet_gpu_sc(
    object = sc_object,
    scrublet_params = gpu_params
  )
})

scrublet_gpu
```

```{r}
#| label: gpu-plot
#| fig-width: 7
#| fig-height: 4
plot(scrublet_gpu)
```

The result is bixverse's `ScrubletRes`, not a GPU-specific class, so
`get_data()`, `call_doublets_manual()` and `add_sc_new_obs()` all work exactly
as they do on the CPU path.

```{r}
#| label: gpu-metrics
gpu_dt <- get_data(scrublet_gpu)
gpu_dt[, Barcode := get_cell_names(sc_object)]
gpu_dt <- merge(gpu_dt, demuxlet_data, by = "Barcode")

doublet_metrics(predicted = gpu_dt$doublet, actual = gpu_dt$Call)
```

## Picking a kNN backend

The kNN over the combined embedding is the one stage where you get a real
choice, and `params_scrublet_gpu()` exposes it through `knn_backend`.

`"gpu"` with `knn_method = "exhaustive"` is the default. It is exact, has no
tuning knobs, and is usually the right answer up to a few hundred thousand
cells. `"ivf"` is approximate, flat in `k`, and only starts paying off above
that. Recall loss matters more here than it does elsewhere, because the doublet
score is a neighbour count: drop neighbours and you bias every score downwards.

```{r}
#| label: knn-backends
#| eval: false
params_ivf <- params_scrublet_gpu(
  expected_doublet_rate = 0.12,
  knn = list(knn_method = "ivf", n_list = 512L, n_probe = 32L)
)

params_cpu_knn <- params_scrublet_gpu(
  expected_doublet_rate = 0.12,
  knn_backend = "cpu",
  knn = list(knn_method = "hnsw")
)
```

`knn_backend = "cpu"` keeps the CPU indices (kmknn, HNSW, Annoy, NN-descent)
while the SVD and the projection stay on the device. It costs a host round trip
on a matrix that is `(1 + sim_doublet_ratio) * n_cells` rows tall, so it is
there for reproducing a CPU run, not for speed.

The two backends take different keys. `knn` is validated against whichever
backend you named, and a key that belongs to the other one errors rather than
being silently ignored:

```{r}
#| label: knn-key-error
#| error: true
params_scrublet_gpu(knn = list(m = 32L))
```

## CPU versus GPU

Same object, same parameters, same seed.

```{r}
#| label: cpu-scrublet
cpu_params <- params_scrublet(
  expected_doublet_rate = 0.12,
  pca = list(no_pcs = gpu_params$no_pcs)
)

cpu_time <- system.time({
  scrublet_cpu <- scrublet_sc(
    object = sc_object,
    scrublet_params = cpu_params
  )
})

data.table(
  version = c("CPU", "GPU"),
  seconds = round(c(cpu_time[["elapsed"]], gpu_time[["elapsed"]]), 2)
)[, speed_up := round(seconds[1] / seconds, 2)][]
```

## Do the calls agree?

```{r}
#| label: agreement
cpu_dt <- get_data(scrublet_cpu)
cpu_dt[, Barcode := get_cell_names(sc_object)]
cpu_dt <- merge(cpu_dt, demuxlet_data, by = "Barcode")

cpu_metrics <- doublet_metrics(cpu_dt$doublet, cpu_dt$Call)
gpu_metrics <- doublet_metrics(gpu_dt$doublet, gpu_dt$Call)

data.table(
  version = c("CPU", "GPU"),
  precision = round(c(cpu_metrics$precision, gpu_metrics$precision), 3),
  recall = round(c(cpu_metrics$recall, gpu_metrics$recall), 3),
  f1 = round(c(cpu_metrics$f1, gpu_metrics$f1), 3),
  threshold = round(c(scrublet_cpu$threshold, scrublet_gpu$threshold), 4)
)
```

```{r}
#| label: score-correlation
cor(
  scrublet_cpu$doublet_scores_obs,
  scrublet_gpu$doublet_scores_obs,
  method = "pearson"
)
```

The scores correlate, they do not match. Both paths use a randomised SVD but
draw a different sketch, and the GPU index breaks neighbour ties differently.
On top of that Otsu's threshold is a step function of the histogram bins, so
the cut can land one bin over and flip a handful of borderline cells. Judge the
two by their metrics, not by a diff.

## Per-sample runs

Doublet rates track cell loading, so a pooled experiment usually wants one
threshold per sample rather than one for everything. `group_by` runs the whole
pipeline per level and stitches the results back together in original cell
order.

```{r}
#| label: grouped
#| eval: false
scrublet_grouped <- scrublet_gpu_sc(
  object = sc_object,
  scrublet_params = gpu_params,
  group_by = "sample_id"
)

scrublet_grouped$threshold
```

`threshold` comes back as a named vector, one entry per group, and
`call_doublets_manual(..., for_sample = "sample_A")` adjusts a single group
without touching the others.

## Clean up

```{r}
#| label: cleanup
unlink(tempdir_doublet, recursive = TRUE, force = TRUE)
```

## What is next

HVG selection is the obvious next lever. It is a full pass over the gene-major
store and on wide data it can outweigh the SVD it feeds. The doublet simulation
is disk-bound rather than compute-bound, so that one probably stays on the host.
Watch this space.
