---
title: "GPU-accelerated SEACells"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{GPU-accelerated SEACells}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Gregor Lueg"
---

## Intro

SEACells ([Persad et al., 2023](https://www.nature.com/articles/s41587-023-01716-9))
finds meta cells by kernel archetypal analysis. The expensive part is the pair
of inner Frank-Wolfe loops: the `B` update scans a gradient over all cells for
every archetype and takes an argmin, and the `A` update solves a column per
cell. Both now run on the GPU in `generate_seacells_gpu_sc()`.

The rest stays on the CPU: the kNN graph, the kernel matrix, the archetype
initialisation, the `K²B` bookkeeping, the RSS evaluation and the aggregation
into pseudo-bulk counts. So the speedup is bounded by how much of the run the
two solves own. That share grows with `n_sea_cells`, because both scale with
`n_cells x k` while most of the CPU-side work does not. Ask for 50 meta cells
and the win is small. Ask for a few thousand and it gets interesting.

Two things worth knowing about the kernels. Each solve picks a workgroup width
from a small set of tiers, and falls back to its CPU sibling for that iteration
if no tier covers your `k`, so a very large `n_sea_cells` can quietly land back
on the host. And nothing on the GPU path holds a dense `n x k` buffer: `B`,
`K²B` and the transposed `A` all stay sparse, which is what makes million-cell
runs fit at all.

If you have not seen the CPU version, read the
[bixverse meta cell vignette](https://gregorlueg.github.io/bixverse/articles/single_cell_meta_cells.html)
first. The params, the result class and everything downstream are identical
here.

```{r}
#| label: setup
library(bixverse)
library(bixverse.gpu)
library(data.table)
library(magrittr)
```

::: {.callout-note}
Vignettes were built locally on a MacBook Pro M1 Max. The GH runners were just
too slow and do not have proper GPU support. This gives an idea of speed on a
decent, but older machine.
:::

## Rebuilding a processed CD34

We use the same CD34 cells from the [SEACells vignette](https://github.com/dpeerlab/SEACells/blob/main/notebooks/SEACell_computation.ipynb) to set up the parent `SingleCells` object: 
load (QC not needed - it's already filtered data), HVG selection, PCA, 
kNN graph. We can use the provided cell type labels to check purity.

```{r}
#| label: rebuild-object
#| code-fold: true
#| code-summary: "Rebuild the CD34 cell object (click to expand)"
cd34_path <- download_cd34_data()

tempdir_cd34 <- tempdir()

sc_object <- SingleCells(dir_data = tempdir_cd34)

sc_object <- load_h5ad(object = sc_object, h5_path = cd34_path)

sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 2000L,
  .verbose = FALSE
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L,
  sparse_svd = FALSE
)

sc_object <- find_neighbours_sc(
  object = sc_object,
  neighbours_params = params_sc_neighbours(
    knn = list(ann_dist = "euclidean", knn_method = "kmknn")
  )
)

# we need the kNN object for the diffusion maps
knn_object <- get_knn_obj(x = sc_object)
```

## Running it

One call, and the params are the same `params_sc_seacells()` the CPU version
takes. We will go with the same parameters as for the other vignette.

```{r}
#| label: gpu-seacells
seacell_params <- params_sc_seacells(
  n_sea_cells = 250L,
  convergence_epsilon = 0.001,
  pruning = TRUE
)

gpu_time <- system.time({
  mc_gpu <- generate_seacells_gpu_sc(
    sc_object,
    seacell_params = seacell_params,
    .verbose = TRUE
  )
})

mc_gpu
```

The result is a `MetaCells` object, byte for byte the same shape as the CPU one.
Assignments live in `@original_assignment`, the RSS trace and the archetype cell
indices in `@other_data`.

```{r}
#| label: gpu-seacells-inspect
head(mc_gpu[[]], 3L)

tail(mc_gpu@other_data$rss, 5L)
```

### Knobs that matter on large data

Two params change the picture once you leave toy datasets:

- **`n_landmarks`** switches archetype initialisation to a Nystroem
  approximation on a landmark subsample. Without it, initialisation is
  quadratic in cells and starts to dominate the run well before the
  Frank-Wolfe loop does.
- **`pruning`** drops tiny values during the Frank-Wolfe updates. It costs a bit
  of numerical fidelity and buys a lot of memory, which is what makes million-
  cell runs feasible at all. `pruning_threshold` controls how aggressive it is.

```{r}
#| label: gpu-seacells-knobs
#| eval: false
seacell_params_large <- params_sc_seacells(
  n_sea_cells = 5000L,
  n_landmarks = 20000L,
  pruning = TRUE,
  pruning_threshold = 1e-7,
  knn = list(k = 25L)
)
```

## CPU versus GPU

Same object, same params, same seed.

```{r}
#| label: cpu-seacells
cpu_time <- system.time({
  mc_cpu <- generate_seacells_sc(
    sc_object,
    seacell_params = seacell_params,
    .verbose = TRUE
  )
})

data.table(
  version = c("CPU", "GPU"),
  seconds = round(c(cpu_time[["elapsed"]], gpu_time[["elapsed"]]), 2)
)[, speed_up := round(seconds[1] / seconds, 2)][]
```

The two do not produce bit-identical assignments. GPU reduction ordering breaks
ties differently, so a cell sitting almost equidistant between two archetypes
can land either way. What should agree is the structure. Compare how the meta
cells partition the cells:

```{r}
#| label: cpu-vs-gpu
mc_sizes <- data.table(
  version = c("CPU", "GPU"),
  n_meta_cells = c(nrow(mc_cpu[[]]), nrow(mc_gpu[[]])),
  median_cells = c(
    median(mc_cpu[[]]$no_originating_cells),
    median(mc_gpu[[]]$no_originating_cells)
  ),
  final_rss = round(
    c(
      tail(mc_cpu@other_data$rss, 1L),
      tail(mc_gpu@other_data$rss, 1L)
    ),
    2
  )
)

mc_sizes
```

We can already see a difference here in speed, and this is a small data set
asking for only 250 meta cells. The more cells and SEACells you throw at the
problem, the bigger the delta: upstream measures the full fit at roughly 2.6x
on 50k cells and 666 archetypes.

## Feeding into SCENIC

The resulting `MetaCells` goes straight into the GPU SCENIC driver, no
conversion step.

```{r}
#| label: metacells-to-scenic
#| eval: false
scenic_res <- scenic_grn_sc_gpu(
  object = mc_gpu,
  tf_ids = tf_dt$gene_id,
  scenic_params = params_scenic(learner_type = "extratrees"),
  .verbose = TRUE
)
```

See the [GPU SCENIC vignette](gpu_scenic.html) for the full workflow.

## What is next

Neither kernel is the bottleneck any more. Each beats the CPU scan it replaces
by several fold in isolation, but the fit as a whole moves by less, and the gap
is Amdahl: the blocking readbacks around the two solves were about a fifth of
wall-clock when last attributed, and the RSS evaluation plus the `K²B`
maintenance still sit on the host. Overlapping the readbacks and moving RSS are
the next two levers. Watch the space.
