---
title: "GPU-accelerated UMAP"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{GPU-accelerated UMAP}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

## Intro

`bixverse.gpu` provides a GPU-accelerated [UMAP](https://arxiv.org/abs/1802.03426)
via `umap_gpu()`. The surface mirrors the CPU version in
[manifoldsR](https://gregorlueg.github.io/manifoldsR/articles/umap.html), so
if you've used that, you already know how to drive this. Kernels run on
[cubecl with the wgpu backend](https://github.com/tracel-ai/cubecl), so any
compatible GPU works (including Apple Silicon).

Current split of work:

- **kNN graph construction**: GPU. Three backends (`"nndescent"`, `"ivf"`,
  `"exhaustive"`), all GPU-accelerated.
- **Fuzzy simplicial set + init** (spectral, PCA, or random): CPU. Porting
  these to GPU is on the roadmap.
- **Embedding optimisation**: GPU. Adam on device (`optimiser = "adam_gpu"`)
  is the default and usually the fastest.

The two heavy hitters (kNN, optimisation) run on the GPU, which is where
most of the wall-clock sits for medium-to-large data. The CPU-side init and
fuzzy-membership steps are comparatively cheap.

If you haven't read the manifoldsR UMAP vignette, that one covers what UMAP
is good at (cluster separation) and where it falls over (continuous
manifolds). Both caveats apply here; only the compute path is different.

```{r}
#| label: setup
library(bixverse.gpu)
library(manifoldsR)
library(data.table)
library(ggplot2)
```

::: {.callout-note}
Vignettes are built on GitHub CI/CD runners with no GPU. GPU code falls back
to software Vulkan (lavapipe), so N is dialled down here for build time. Real
hardware handles much bigger N without issue. Also, there are some weird 
problems on the GH runners with Lavapipe compiling the wgpu code that affects
specifically the IVF indices. `eval = FALSE` for these code chunks.
:::

## Generating data

Focus is on the case UMAP handles well: clustered data. 25k points across
25 clusters in 32D. Small enough to run fast, big enough that the GPU
optimiser has something to chew on.

```{r}
#| label: data
set.seed(42L)

cluster_data <- manifold_synthetic_data(
  type = "clusters",
  n_samples = 10000L,
  dim = 32L,
  parameters = params_clusters(n_clusters = 25L)
)
```

## Running UMAP on GPU

Default call: NN-descent on GPU for kNN, spectral init on CPU, GPU Adam for
the embedding.

```{r}
#| label: umap default
umap_default <- umap_gpu(
  data = cluster_data$data,
  k = 15L,
  min_dist = 0.5,
  spread = 1.0,
  knn_method = "nndescent",
  umap_params = params_umap_gpu(),
  seed = 42L,
  .verbose = FALSE
)

plot_df <- as.data.table(umap_default) |>
  setnames(c("UMAP1", "UMAP2"))
plot_df[, cluster := as.factor(cluster_data$membership)]

ggplot(plot_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(colour = cluster), size = 0.5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("umap_gpu, default (NN-descent + adam_gpu)")
```

Clusters separate cleanly.

## Choosing a kNN backend

Three GPU backends via `knn_method`:

- **`"nndescent"`**: NN-descent with CAGRA-style graph pruning. Solid default.
- **`"ivf"`**: inverted file index over Voronoi cells. Wins on large data
  where an approximate answer is fine.
- **`"exhaustive"`**: exact brute force. Small data or ground-truth checks.
  Quadratic in N, gets painful fast.

Tuning knobs live in `params_nn_gpu()`, defaults are fine for most cases.

### IVF backend

```{r}
#| label: umap ivf
#| eval: false

# not evaluated on CI: lavapipe/cubecl miscompile, runs fine on real GPUs
umap_ivf <- umap_gpu(
  data = cluster_data$data,
  k = 15L,
  knn_method = "ivf",
  umap_params = params_umap_gpu(),
  seed = 42L,
  .verbose = FALSE
)

plot_df_ivf <- as.data.table(umap_ivf) |>
  setnames(c("UMAP1", "UMAP2"))
plot_df_ivf[, cluster := as.factor(cluster_data$membership)]

ggplot(plot_df_ivf, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(colour = cluster), size = 0.5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("umap_gpu, IVF kNN")
```

### Exhaustive backend

```{r}
#| label: umap exhaustive
umap_exhaustive <- umap_gpu(
  data = cluster_data$data,
  k = 15L,
  knn_method = "exhaustive",
  umap_params = params_umap_gpu(),
  seed = 42L,
  .verbose = FALSE
)

plot_df_ex <- as.data.table(umap_exhaustive) |>
  setnames(c("UMAP1", "UMAP2"))
plot_df_ex[, cluster := as.factor(cluster_data$membership)]

ggplot(plot_df_ex, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(colour = cluster), size = 0.5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("umap_gpu, exhaustive kNN")
```

Structurally the three embeddings agree. Differences you see are within
the noise of UMAP's non-deterministic optimisation. The approximate methods
are indistinguishable from exact for downstream analysis.

## Choosing an optimiser

Four optimisers via `params_umap_gpu(optimiser = ...)`:

- **`"adam_gpu"`**: GPU Adam, full optimisation on device with parallel
  gradient collection. Default, usually fastest.
- **`"adam_parallel"`**: CPU parallel Adam (the fast `manifoldsR` default).
  Handy if you want to keep the optimiser on CPU while still using the GPU
  kNN backend.
- **`"adam"`**: sequential CPU Adam. Can produce slightly cleaner embeddings
  on small data.
- **`"sgd"`**: classic stochastic gradient descent, as in the original UMAP
  paper. Slower, worth trying on continuous manifolds where you want a low
  `min_dist` to shine.

### SGD optimiser

```{r}
#| label: umap sgd
umap_sgd <- umap_gpu(
  data = cluster_data$data,
  k = 15L,
  min_dist = 0.3, # SGD prefers smaller min_dist
  knn_method = "nndescent",
  umap_params = params_umap_gpu(optimiser = "sgd"),
  seed = 42L,
  .verbose = FALSE
)

plot_df_sgd <- as.data.table(umap_sgd) |>
  setnames(c("UMAP1", "UMAP2"))
plot_df_sgd[, cluster := as.factor(cluster_data$membership)]

ggplot(plot_df_sgd, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(colour = cluster), size = 0.5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("umap_gpu, SGD optimiser")
```

### Parallel Adam optimiser

CPU parallel Adam (the `manifoldsR::umap()` default). Useful as a reference
or when you want the optimiser on CPU and the kNN on GPU.

```{r}
#| label: umap adam parallel
umap_ap <- umap_gpu(
  data = cluster_data$data,
  k = 15L,
  min_dist = 0.5,
  knn_method = "nndescent",
  umap_params = params_umap_gpu(optimiser = "adam_parallel"),
  seed = 42L,
  .verbose = FALSE
)

plot_df_ap <- as.data.table(umap_ap) |>
  setnames(c("UMAP1", "UMAP2"))
plot_df_ap[, cluster := as.factor(cluster_data$membership)]

ggplot(plot_df_ap, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(colour = cluster), size = 0.5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("umap_gpu, adam_parallel optimiser")
```

## Using a pre-computed kNN graph

Already have a `NearestNeighbours` object (from `manifoldsR` or from one of
the kNN routines in this package)? Hand it to `umap_gpu()` via the `knn`
argument and skip the graph build. Handy when sweeping UMAP parameters.

```{r}
#| label: umap precomputed knn
knn_precomputed <- generate_knn_graph_gpu(
  data = cluster_data$data,
  k = 15L,
  .verbose = FALSE
)

umap_from_knn <- umap_gpu(
  data = cluster_data$data,
  knn = knn_precomputed,
  k = 15L,
  min_dist = 0.5,
  umap_params = params_umap_gpu(),
  seed = 42L,
  .verbose = FALSE
)

plot_df_knn <- as.data.table(umap_from_knn) |>
  setnames(c("UMAP1", "UMAP2"))
plot_df_knn[, cluster := as.factor(cluster_data$membership)]

ggplot(plot_df_knn, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(colour = cluster), size = 0.5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("umap_gpu, pre-computed kNN")
```

Sweeping `min_dist` is essentially free now; the expensive kNN build only
runs once:

```{r}
#| label: umap min dist sweep
umap_wide <- umap_gpu(
  data = cluster_data$data,
  knn = knn_precomputed,
  k = 15L,
  min_dist = 1.0,
  umap_params = params_umap_gpu(),
  seed = 42L,
  .verbose = FALSE
)

plot_df_wide <- as.data.table(umap_wide) |>
  setnames(c("UMAP1", "UMAP2"))
plot_df_wide[, cluster := as.factor(cluster_data$membership)]

ggplot(plot_df_wide, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(colour = cluster), size = 0.5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("umap_gpu, min_dist = 1.0")
```

## When does the GPU version pay off?

The CPU version in `manifoldsR` is already fast: parallel Adam, faer under
the hood, full SIMD kNN toolkit. On Apple Silicon it's especially
competitive because wgpu doesn't expose proper tensor-core equivalents on
that hardware. On tall matrices with many samples the GPU optimiser and GPU
kNN clearly pull ahead. On smaller data, kernel launch and host-device
transfer overhead can wipe out the win.

Rough shape:

- Below ~10k points: `manifoldsR`'s CPU parallel Adam is competitive or
  faster. GPU pays a fixed setup cost.
- 10k to 100k: GPU pulls ahead, especially with `"ivf"` or `"nndescent"`.
- Above 100k: GPU clearly faster, scales more gracefully with N.

Other axis is the kNN backend. `"exhaustive"` is quadratic in N, keep it
for smaller data or ground-truth checks. `"ivf"` and `"nndescent"` are the
workhorses.

## Conclusions

Same mental model as CPU UMAP, same knobs, same caveats (UMAP is a
visualisation tool, not a metric space!). Current pipeline is a hybrid:
kNN and optimisation on GPU, fuzzy membership and init still on CPU. Both
CPU steps are cheap and the GPU parts already deliver most of the speedup,
but they'll move to GPU eventually.

Downstream this ties into the single cell workflow: PCA on GPU, Harmony v2
on GPU, kNN on GPU, then UMAP on GPU. Minimum number of CPU handoffs in the
reduce-embed-visualise chain.
