---
title: "GPU-accelerated workflows for bixverse single cell"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{GPU-accelerated workflows in bixverse single cell}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Gregor Lueg"
---

## Intro

This vignette walks through a GPU-accelerated single cell workflow on the
two-batch PBMC data set (pbmc3k + pbmc4k) used in the
[batch correction vignette](https://gregorlueg.github.io/bixverse/articles/single_cell_batch_correction.html).
The point is to demonstrate how `bixverse.gpu` can drop into the heavier steps
of a standard pipeline: PCA, Harmony v2 batch correction, and the kNN search.
Two batches give us a natural reason to run all three in sequence: PCA
generates the embedding, Harmony corrects it, and the kNN methods build the
neighbour graph for downstream clustering. If none of this single cell stuff
makes sense in the `bixverse` framework, please read
[this](https://gregorlueg.github.io/bixverse/articles/thinking_single_cell.html)
first.

The core idea behind `bixverse.gpu` is hardware-agnostic GPU code, hence the
use of [cubecl with the wpgu backend](https://github.com/tracel-ai/cubecl).
For small data sets the host-to-device transfer and kernel launch overhead can
outweigh the speedup, but the larger the data, the more these methods pay off
(if you have sufficient VRAM/unified memory that is...).

`bixverse.gpu` currently provides:

- **GPU PCA** via `calculate_pca_gpu_sc()`: a sparse, randomised SVD where the
  large matrix multiplications run on GPU. Scaling (if desired) is applied 
  without ever materialising the dense matrix. Also, option for the PFLogPF
  normalisation.
- **GPU Harmony v2** via `harmony_v2_gpu_sc()`: a GPU implementation of
  Harmony v2 ([Patikas et al., 2026](https://www.biorxiv.org/content/10.64898/2026.03.16.711825v1))
  with the Arrowhead matrix inversion. Single batch covariate only on the GPU
  path. R is refined via full-batch Jacobi sweeps rather than the blockwise 
  updates of the original: faster, but results are very similar rather than 
  bit-identical. GPU floating point reduction ordering adds small
  further deviations.
- **GPU kNN** via two generics:
  - `find_neighbours_gpu_sc()` covers **exhaustive** (exact brute force) and
    **IVF** (inverted file index, approximate).
  - `find_neighbours_cagra_sc()` covers **CAGRA**, which prunes an NNDescent
    graph and supports either direct kNN extraction or beam search.

```{r}
#| label: setup
library(bixverse)
library(bixverse.gpu)
library(bixverse.plots)
library(data.table)
library(ggplot2)
```

::: {.callout-note}
Vignettes were built locally on a MacBook Pro M1 Max. The GH runners were just
too slow and do not have proper GPU support. This gives an idea of speed on a 
decent, but older machine.
:::

## Loading the data

The pbmc3k and pbmc4k data sets share tissue but differ in sequencing depth
and cell counts, which produces a clear batch effect.

```{r}
#| label: load data
#| code-fold: true
dir_data <- download_pbmc_batches()

tempdir_gpu <- file.path(tempdir(), "gpu_workflow")
dir.create(tempdir_gpu, showWarnings = FALSE, recursive = TRUE)

h5ad_files <- list.files(dir_data)
h5ad_files <- h5ad_files[grepl(".h5ad", h5ad_files)]
h5ad_paths <- file.path(dir_data, h5ad_files)
names(h5ad_paths) <- c("pbmc3k", "pbmc4k")

h5_tasks <- prescan_h5ad_files(h5_paths = h5ad_paths)

sc_object <- SingleCells(dir_data = tempdir_gpu)

sc_object <- load_multi_h5ad(
  object = sc_object,
  prescan_result = h5_tasks,
  .verbose = TRUE
)
```

## Quality control

Standard QC -> mitochondrial proportions, library size and complexity outliers.

```{r}
#| label: qc
#| code-fold: true
var <- get_sc_var(sc_object)

h5_metadata <- read_h5ad_metadata(h5ad_paths[[1]])

var <- merge(
  var,
  h5_metadata$var[, c("ENSEMBL_ID", "Symbol_TENx")],
  by.x = "gene_id",
  by.y = "ENSEMBL_ID"
)

setnames(var, old = "Symbol_TENx", new = "gene_symbol", skip_absent = TRUE)

gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_symbol), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_symbol), gene_id]
)

sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  streaming = FALSE,
  .verbose = TRUE
)

qc_df <- sc_object[[c("cell_id", "lib_size", "nnz", "MT")]]

metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)
directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)

qc <- run_cell_qc(
  metrics = metrics,
  cells_to_keep = get_cells_to_keep(sc_object),
  directions = directions,
  threshold = 3
)

sc_object[["outlier"]] <- qc$combined
cells_to_keep <- qc_df[!qc$combined, cell_id]
sc_object <- set_cells_to_keep(sc_object, cells_to_keep)
```

## Highly variable genes

HVG selection stays on the CPU.

```{r}
#| label: hvg
sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 2000L,
  .verbose = TRUE
)
```

## GPU-accelerated PCA

However, with PCA things start changing. Let's run first the CPU version which
you know

```{r}
#| label: pca cpu 
sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 32L,
  sparse_svd = TRUE
)
# extract the factors and singular values
cpu_res <- get_pca_factors(sc_object)
cpu_s <- get_pca_singular_val(sc_object)
```

The GPU version looks very similar

```{r}
#| label: gpu pca
sc_object <- calculate_pca_gpu_sc(object = sc_object, no_pcs = 32L)
gpu_res <- get_pca_factors(sc_object)
gpu_s <- get_pca_singular_val(sc_object)
``` 

Let's compare against the CPU version: 
```{r}
#| label: cpu vs gpu - pc1
ggplot(
  data = data.table(PC1_cpu = cpu_res[, 1], PC1_gpu = gpu_res[, 1]),
  mapping = aes(x = PC1_cpu, y = PC1_gpu)
) +
  geom_point() +
  theme_bw() +
  xlab("PC1 (CPU)") +
  ylab("PC1 (GPU)") +
  ggtitle("CPU vs GPU")
``` 

```{r}
#| label: cpu vs gpu - singular values
ggplot(
  data = data.table(sv_cpu = cpu_s, sv_gpu = gpu_s),
  mapping = aes(x = sv_cpu, y = sv_gpu)
) +
  geom_point() +
  theme_bw() +
  xlab("Singular values (CPU)") +
  ylab("Singular values (GPU)") +
  ggtitle("CPU vs GPU")
```

You might see some slight differences here driven by floating operation 
differences between CPU and GPU. The overall data structure is however clearly
captured.

## GPU-accelerated Harmony v2

With two batches we need batch correction. `harmony_v2_gpu_sc()` runs Harmony
v2 with the Arrowhead inversion on GPU and writes the corrected embedding to
the object as `"harmony_gpu"`.

```{r}
#| label: harmony gpu
sc_object <- harmony_v2_gpu_sc(
  object = sc_object,
  batch_column = "exp_id",
  harmony_params = params_sc_harmony_v2_gpu()
)
```

For comparison, the CPU version of v2; stored as `"harmony_v2"` so both
embeddings coexist.

```{r}
#| label: harmony cpu
sc_object <- harmony_v2_sc(
  object = sc_object,
  batch_column = "exp_id",
  harmony_params = params_sc_harmony_v2()
)
```

We will compare the two below, but only once we have kNN graphs on each
embedding, since most of the batch correction metrics need a neighbourhood
structure to begin with. Which is a convenient segue.

## GPU-accelerated kNN

Three methods, each with different speed/precision trade-offs. Below we run
all three against the GPU Harmony embedding to show the API. In practice, pick
one based on data size and how exact you need the neighbours to be.

### Exhaustive

Exact brute-force search on GPU. Best for smaller data sets or whenever you
need exact neighbours. Scales quadratically, so it gets painful on large data.

```{r}
#| label: gpu exhaustive
sc_object <- find_neighbours_gpu_sc(
  object = sc_object,
  embd_to_use = "harmony_gpu",
  gpu_method = "exhaustive",
  k = 15L,
  dist_metric = "euclidean",
  .verbose = TRUE
)
```

### IVF

Inverted file index. Partitions the embedding space into Voronoi cells and
probes only a subset at query time. Worthwhile on larger data sets. The key
knobs (`nlist` and `nprobe`) live in `params_sc_ivf()`; defaults are fine
here.

```{r}
#| label: gpu ivf
sc_object <- find_neighbours_gpu_sc(
  object = sc_object,
  embd_to_use = "harmony_gpu",
  gpu_method = "ivf",
  ivf_params = params_sc_ivf(),
  .verbose = TRUE
)
```

### CAGRA

Builds a pruned NNDescent graph, based on the fantastic work by [Nvidia](https://arxiv.org/pdf/2308.15136). (Unfortunately, some of the cuda primitives are not available
in wgpu, but it is still a blazingly fast approximate kNN search even on wgpu).
With `extract_knn = TRUE` you pull the kNN straight out of the NNDescent graph: 
faster, slightly less precise, useful for rapid iteration over parameters. 
With `extract_knn = FALSE` (the default) the function runs beam search over the 
pruned graph for higher recall, which matters more on larger, higher-dimensional 
data.

```{r}
#| label: gpu cagra
sc_object <- find_neighbours_cagra_sc(
  object = sc_object,
  embd_to_use = "harmony_gpu",
  cagra_params = params_sc_cagra(),
  extract_knn = FALSE,
  .verbose = TRUE
)
```

From here downstream methods, clustering, UMAP/tSNE, marker detection,
work without modification, exactly as after a `find_neighbours_sc()` call.

## Comparing GPU vs CPU Harmony

With the kNN graph in place on the GPU Harmony embedding, we can compute batch
metrics for it and then repeat the kNN step on the CPU embedding to compare.

```{r}
#| label: gpu harmony metrics
kbet_gpu <- calculate_kbet_sc(sc_object, batch_column = "exp_id")
asw_gpu <- calculate_batch_asw_sc(
  sc_object,
  embd_to_use = "harmony_gpu",
  batch_column = "exp_id"
)
lisi_gpu <- calculate_batch_lisi_sc(sc_object, batch_column = "exp_id")

kbet_gpu
asw_gpu
lisi_gpu
```

Same kNN setup on the CPU Harmony embedding:

```{r}
#| label: cpu harmony knn + metrics
sc_object <- find_neighbours_cagra_sc(
  object = sc_object,
  embd_to_use = "harmony_v2",
  cagra_params = params_sc_cagra(),
  extract_knn = FALSE,
  .verbose = TRUE
)

kbet_cpu <- calculate_kbet_sc(sc_object, batch_column = "exp_id")
asw_cpu <- calculate_batch_asw_sc(
  sc_object,
  embd_to_use = "harmony_v2",
  batch_column = "exp_id"
)
lisi_cpu <- calculate_batch_lisi_sc(sc_object, batch_column = "exp_id")

kbet_cpu
asw_cpu
lisi_cpu
```

Harmony has stochastic elements, so the two embeddings will not be identical,
but the batch correction quality should be in the same ballpark across the
metrics.

### UMAP on the GPU Harmony embedding

Everyone loves visuals, even if you should not over-interpret them
([Chari et al., 2023](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1011288)).
`umap_gpu_sc()` runs the full GPU-accelerated UMAP path: GPU kNN plus a
GPU Adam optimiser (`optimiser = "adam_gpu"` in `params_umap_gpu()`, the
default). On the CAGRA kNN we just computed above, it plugs straight in
with `use_knn = TRUE`.

```{r}
#| label: umap gpu harmony
sc_object <- find_neighbours_cagra_sc(
  object = sc_object,
  embd_to_use = "harmony_gpu",
  cagra_params = params_sc_cagra(),
  extract_knn = FALSE,
  .verbose = TRUE
)

sc_object <- umap_gpu_sc(
  object = sc_object,
  embd_to_use = "harmony_gpu",
  slot_name = "umap_harm_gpu",
  use_knn = TRUE
)
```

```{r}
#| label: umap plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "umap_harm_gpu",
  colour_by = "exp_id",
  discrete = TRUE
) +
  labs(
    title = "GPU Harmony v2 + GPU CAGRA kNN + GPU UMAP",
    colour = "Batch:"
  )
```

### tSNE on the GPU Harmony embedding

`tsne_gpu_sc()` mirrors the shape of `umap_gpu_sc()` but only the kNN step
runs on GPU. The optimiser (Barnes-Hut or FIt-SNE FFT) still runs on CPU;
a GPU optimiser is on the roadmap. Because t-SNE derives `k` from
`3 * perplexity` on the Rust side, `use_knn` defaults to `FALSE` so that
every call builds a fresh GPU kNN sized to the requested perplexity.
Handy when sweeping perplexities.

```{r}
#| label: tsne gpu harmony
sc_object <- tsne_gpu_sc(
  object = sc_object,
  embd_to_use = "harmony_gpu",
  slot_name = "tsne_harm_gpu",
  perplexity = 30.0
)
```

```{r}
#| label: tsne plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "tsne_harm_gpu",
  colour_by = "exp_id",
  discrete = TRUE
) +
  labs(
    title = "GPU Harmony v2 + GPU t-SNE (GPU kNN, CPU optimiser)",
    colour = "Batch:"
  )
```

## Conclusions

The full GPU path (PCA, Harmony v2, kNN, UMAP with GPU Adam optimiser,
t-SNE with GPU kNN) plugs into the existing `SingleCells` workflow without
any glue code. The downstream object behaves identically to whatever you
would get from the CPU equivalents.

Longer term, the GPU kNN methods are the obvious building block for
GPU-accelerated versions of methods that lean heavily on kNN graphs:

- **BBKNN** ([paper](https://academic.oup.com/bioinformatics/article/36/3/964/5545955),
  CPU [implementation](https://gregorlueg.github.io/bixverse/reference/bbknn_sc.html))
- **fastMNN** ([paper](https://www.nature.com/articles/nbt.4091),
  CPU [implementation](https://gregorlueg.github.io/bixverse/reference/fast_mnn_sc.html))

Doublet detection is another candidate. Another potentially interesting area
would be GPU-accelerated NMF ... ? Basically, watch the space.

## Clean up

```{r}
#| label: clean up
unlink(tempdir_gpu, recursive = TRUE, force = TRUE)
```
