---
title: "GPU-accelerated NMF"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{GPU-accelerated NMF}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Gregor Lueg"
---

## Intro

NMF factorises the counts into `V ≈ W H` with everything non-negative, which is
why the components read as additive gene programmes rather than the
plus-and-minus directions PCA gives you. `bixverse` solves it with HALS
(hierarchical alternating least squares), and this package moves that solver
onto the GPU.

What actually runs on the device: both Gram products, both data products, the
two HALS sweeps, the column normalisation and the objective. `V`, `W` and `H`
stay resident, so the only thing crossing back per iteration is a small
partials vector.

What does not: the NNDSVD initialisation, and the whole consensus machinery.
Pooling the components across restarts, the local-density filter, the k-means,
the silhouette and the per-coordinate median all run on the CPU, shared verbatim
with `bixverse`. That is deliberate. Those operate on a `(k * n_runs) x dim`
matrix with `k * n_runs` in the hundreds, so they are nowhere near the solve in
cost and porting them would buy nothing.

So where does the GPU actually help? Not much in a single fit. It helps when the
same matrix serves many solves, because `V` uploads once and every restart and
every rank reuses it. On the CPU each of those solves pays full memory traffic
over `V` again. The k sweep is `length(k_range) * n_runs` solves over one
matrix, which is exactly that shape. On the 6.9k cells below, 20 restarts come
back a bit over 3x faster.

One limit to know up front: the kernels tier their workgroup width by rank and
stop at 128. Ask for more and you get an error pointing you at the CPU version,
not a silent fallback.

The params, the result classes and everything downstream are identical to the
CPU versions. If you have not seen those, read the
[bixverse bag-of-genes vignette](https://gregorlueg.github.io/bixverse/articles/bag_of_genes_single_cells.html)
first.

```{r}
#| label: setup
library(bixverse)
library(bixverse.gpu)
library(data.table)
library(ggplot2)
library(magrittr)
```

::: {.callout-note}
Vignettes were built locally on a MacBook Pro M1 Max. The GH runners were just
too slow and do not have proper GPU support. This gives an idea of speed on a
decent, but older machine.
:::

## Rebuilding a processed CD34

Same CD34 cells as the other single cell vignettes. Load, pick HVGs, PCA. No QC
needed, the data is already filtered.

```{r}
#| label: rebuild-object
#| code-fold: true
#| code-summary: "Rebuild the CD34 cell object (click to expand)"
cd34_path <- download_cd34_data()

tempdir_cd34 <- tempdir()

sc_object <- SingleCells(dir_data = tempdir_cd34)

sc_object <- load_h5ad(object = sc_object, h5_path = cd34_path)

sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 2000L,
  .verbose = FALSE
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L,
  sparse_svd = FALSE
)

sc_object
```

## Picking k

You do not know `k` in advance, and NMF is non-convex, so a single fit at a
guessed rank tells you very little. Consensus NMF ([Kotliar et al., 2019](https://elifesciences.org/articles/43803))
answers this by running the fit many times from different random starts and
asking which programmes keep coming back.

`nmf_k_sweep_gpu_sc()` does that across a range of ranks and reports stability
(mean silhouette of the consensus clusters) against reconstruction error. It
keeps no factors, so a wide range stays cheap in memory.

```{r}
#| label: nmf-k-sweep
sweep_time <- system.time({
  k_sweep <- nmf_k_sweep_gpu_sc(
    object = sc_object,
    k_range = 2:12,
    n_runs = 10L,
    nmf_consensus_params = params_nmf_consensus(density_threshold = 2)
  )
})

k_sweep
```

```{r}
#| label: nmf-k-sweep-plot
#| fig-cap: "Consensus NMF k sweep on the CD34 cells."
plot(k_sweep)
```

The rule of thumb is to take the last `k` before stability falls away, while
the error curve is still coming down. Error always falls with `k`, so it never
picks a rank on its own; stability is what stops you.

Do not just read off the most stable row. Low ranks are trivially stable, since
two components are hard to disagree about, and the header will happily tell you
`k = 2`. What you want is the last rank that still holds up. Here stability sits
near 1 through `k = 4`, dips at 5, recovers at 6 and 7, then falls away from 8
onward and does not come back. Error is still dropping at 7, so that is where we
fit.

A note on `density_threshold = 2`. The filter drops components sitting in a
sparse neighbourhood, and 2 is the cosine ceiling, so that switches it off. With
10 restarts it has very little to work with, and a run that drops below `k`
survivors errors out rather than quietly returning something worse. Turn it back
on once you are running 50 or more restarts.

## Fitting at the chosen k

```{r}
#| label: consensus-nmf
nmf_res <- consensus_nmf_gpu_sc(
  object = sc_object,
  k = 7L,
  n_runs = 20L,
  nmf_consensus_params = params_nmf_consensus(density_threshold = 2)
)

nmf_res
```

`get_stability()` gives you the diagnostics behind the fit: the mean silhouette,
the relative reconstruction errors, and a row per pooled component recording
where it landed.

```{r}
#| label: consensus-nmf-diagnostics
nmf_diag <- get_stability(nmf_res)

nmf_diag$stability

nmf_diag$cluster_sizes
```

With 20 restarts, a cluster of 20 is a programme every run found. A thin one is
a programme only some initialisations saw, and it is worth a squint before you
build a story on it.

To read off what a programme is, rank the genes by their loading in `W`:

```{r}
#| label: consensus-nmf-top-genes
w_mat <- get_w(nmf_res)

top_genes <- lapply(colnames(w_mat), \(comp) {
  names(sort(w_mat[, comp], decreasing = TRUE))[1:10]
})
names(top_genes) <- colnames(w_mat)

top_genes[1:3]
```

`H` is `k x cells`, so the activations drop straight onto an embedding or into
the obs table.

```{r}
#| label: consensus-nmf-activity
h_mat <- get_h(nmf_res)

dim(h_mat)

round(h_mat[1:3, 1:4], 3)
```

## CPU versus GPU

The honest comparison. Both solvers run the same HALS, the same number of
restarts and the same host-side consensus step, so the only thing that differs
is where the linear algebra happens.

```{r}
#| label: benchmark-consensus
gpu_time <- system.time({
  gpu_fit <- consensus_nmf_gpu_sc(
    object = sc_object,
    k = 7L,
    n_runs = 20L,
    nmf_consensus_params = params_nmf_consensus(density_threshold = 2),
    seed = 42L,
    .verbose = FALSE
  )
})

cpu_time <- system.time({
  cpu_fit <- consensus_nmf_sc(
    object = sc_object,
    k = 7L,
    n_runs = 20L,
    nmf_consensus_params = params_nmf_consensus(density_threshold = 2),
    seed = 42L,
    .verbose = FALSE
  )
})

data.table(
  version = c("CPU", "GPU"),
  seconds = round(c(cpu_time[["elapsed"]], gpu_time[["elapsed"]]), 2)
)[, speed_up := round(seconds[1] / seconds, 2)][]
```

Do the two agree? Not bit for bit, and they cannot: f32 GEMM on the device
reduces in a different order than the CPU path, and NMF is non-convex, so tiny
differences early can send a restart to a different local optimum. What has to
match is the answer.

```{r}
#| label: benchmark-agreement
data.table(
  version = c("CPU", "GPU"),
  rel_error = round(c(cpu_fit$rel_error, gpu_fit$rel_error), 5),
  stability = round(c(cpu_fit$stability, gpu_fit$stability), 4)
)
```

One thing to keep in mind about the restarts: on the CPU they run across cores,
on the GPU they run one after the other on a single device. So the GPU is giving
up the easiest parallelism there is and still has to win on the solve itself.
That is also why the k sweep flatters it more than a single consensus fit does:
more solves over the same resident matrix, and no extra upload for any of them.

## Meta cells

Everything above dispatches on `MetaCells` too, with the same generics. Meta
cells are the path where consensus NMF is genuinely affordable, since a few
thousand aggregates instead of a few hundred thousand cells makes `n_runs = 50`
unremarkable. See the [SEACells vignette](gpu_metacells.html) for getting there.

```{r}
#| label: metacells-nmf
#| eval: false
mc_sweep <- nmf_k_sweep_gpu_sc(
  object = mc_object,
  k_range = 2:12,
  n_runs = 20L
)

mc_nmf <- consensus_nmf_gpu_sc(
  object = mc_object,
  k = 6L,
  n_runs = 20L
)
```

## What is next

The consensus step is now the floor on how fast this can get. Once the solves
move to the device, the pooling and the k-means start showing up in the profile
at large `n_runs`, and clustering in cell space (`consensus_target = "w"`) is
the pathological case: it pools a dense `(k * n_runs) x n_cells` matrix and runs
an exhaustive cosine search over it. That one is worth porting.

The rank cap at 128 is a workgroup-tiering limit rather than anything
fundamental. It has not bitten in practice, since a 128-programme
factorisation of single cell data is well past interpretable, but it is there.
