---
title: "GPU-accelerated t-SNE"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{GPU-accelerated t-SNE}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

## Intro

`bixverse.gpu` provides a t-SNE wrapper via `tsne_gpu()` where the usually
time-consuming kNN step runs on the GPU. The optimiser itself (Barnes-Hut
or FFT-accelerated interpolation) is still on the CPU via the fast Rust
implementation in
[manifoldsR](https://gregorlueg.github.io/manifoldsR/articles/tsne.html).

If you want the algorithm background (perplexity, exaggeration schedules,
BH vs FFT, when to reach for t-SNE over UMAP), read the
[manifoldsR t-SNE vignette](https://gregorlueg.github.io/manifoldsR/articles/tsne.html).
This vignette only covers what's different here: the kNN runs on the GPU.

Current split of work:

- **kNN graph construction**: GPU. Three backends (`"nndescent"`, `"ivf"`,
  `"exhaustive"`).
- **Affinity computation and gradient descent** (BH or FFT): CPU. Same fast
  Rust implementation as `manifoldsR::tsne()`. A GPU optimiser is on the
  roadmap.

For medium-sized data the kNN is usually the dominant cost, so pushing it to
the GPU is the main win. On very large data the optimiser starts to dominate
and the GPU advantage narrows until the optimiser also moves over.

```{r}
#| label: setup
library(bixverse.gpu)
library(manifoldsR)
library(data.table)
library(ggplot2)
```

## Generating data

Same clustered synthetic data as in the UMAP vignette.

```{r}
#| label: data
set.seed(42L)

cluster_data <- manifold_synthetic_data(
  type = "clusters",
  n_samples = 10000L,
  dim = 32L,
  parameters = params_clusters(n_clusters = 25L)
)
```

## Running t-SNE on GPU

Default call: NN-descent on GPU for kNN, PCA init, Barnes-Hut optimiser on
CPU.

```{r}
#| label: tsne default
tsne_default <- tsne_gpu(
  data = cluster_data$data,
  perplexity = 10,
  knn_method = "nndescent",
  tsne_params = params_tsne_gpu(),
  seed = 42L,
  .verbose = TRUE
)

plot_df <- as.data.table(tsne_default) |>
  setnames(c("tSNE1", "tSNE2"))
plot_df[, cluster := as.factor(cluster_data$membership)]

ggplot(plot_df, aes(x = tSNE1, y = tSNE2)) +
  geom_point(aes(colour = cluster), size = 0.5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("tsne_gpu, default (NN-descent + BH)")
```

Clusters separate cleanly, as expected from t-SNE.

## Choosing a kNN backend

Three GPU backends via `knn_method`:

- **`"exhaustive"`**: exact brute force. Small data or ground-truth checks.
  Quadratic in N.
- **`"ivf"`**: inverted file index over Voronoi cells. Wins on large data
  where an approximate answer is fine. Works very well on strongly clustered
  data.
- **`"nndescent"`**: NN-descent with CAGRA-style graph pruning. Solid default.

Tuning knobs live in `params_nn_gpu()`, defaults are fine for most cases.
The knobs are identical to the ones described in the UMAP vignette.

### IVF backend

```{r}
#| label: tsne ivf
tsne_ivf <- tsne_gpu(
  data = cluster_data$data,
  perplexity = 10,
  knn_method = "ivf",
  tsne_params = params_tsne_gpu(),
  seed = 42L,
  .verbose = TRUE
)

plot_df_ivf <- as.data.table(tsne_ivf) |>
  setnames(c("tSNE1", "tSNE2"))
plot_df_ivf[, cluster := as.factor(cluster_data$membership)]

ggplot(plot_df_ivf, aes(x = tSNE1, y = tSNE2)) +
  geom_point(aes(colour = cluster), size = 0.5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("tsne_gpu, IVF kNN")
```

Structurally the embeddings agree. Differences are within the noise of
t-SNE's non-deterministic optimisation.

## FFT-accelerated t-SNE

For large data the FFT variant reduces the optimiser cost from `O(N log N)`
to `O(N)` at the price of a larger constant. Interesting at ≥ 100k samples.
FFT is Unix-only (FFTW cross-compilation constraint).

```{r}
#| label: tsne fft
#| eval: !expr .Platform$OS.type == "unix"
tsne_fft <- tsne_gpu(
  data = cluster_data$data,
  perplexity = 10,
  approx_type = "fft",
  knn_method = "nndescent",
  tsne_params = params_tsne_gpu(),
  seed = 42L,
  .verbose = TRUE
)

plot_df_fft <- as.data.table(tsne_fft) |>
  setnames(c("tSNE1", "tSNE2"))
plot_df_fft[, cluster := as.factor(cluster_data$membership)]

ggplot(plot_df_fft, aes(x = tSNE1, y = tSNE2)) +
  geom_point(aes(colour = cluster), size = 0.5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("tsne_gpu, FFT")
```

If you see the FFT embedding relax into a gapless disc rather than distinct
cluster islands on very large data, that's the equilibrium at exaggeration
= 1. Set `late_exag_factor` in `params_tsne_gpu()` to a value between 2 and
4 to pull the clusters apart. Same caveat as the manifoldsR vignette.

## Using a pre-computed kNN graph

Sweeping t-SNE hyperparameters (perplexity, learning rate, exaggeration) is
usually where you spend the most time. Cache the kNN once and hand it in.

```{r}
#| label: tsne precomputed knn
knn_precomputed <- generate_knn_graph_gpu(
  data = cluster_data$data,
  k = 60L, # ~ 3 * perplexity for perplexity = 30
  knn_method = "ivf",
  .verbose = FALSE
)

tsne_from_knn <- tsne_gpu(
  data = cluster_data$data,
  knn = knn_precomputed,
  perplexity = 30,
  tsne_params = params_tsne_gpu(),
  seed = 42L,
  .verbose = FALSE
)

plot_df_knn <- as.data.table(tsne_from_knn) |>
  setnames(c("tSNE1", "tSNE2"))
plot_df_knn[, cluster := as.factor(cluster_data$membership)]

ggplot(plot_df_knn, aes(x = tSNE1, y = tSNE2)) +
  geom_point(aes(colour = cluster), size = 0.5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("tsne_gpu, pre-computed kNN")
```

Note the kNN needs enough neighbours for the perplexity you're using;
`k ≈ 3 * perplexity` is the usual rule.

## When does the GPU version pay off?

The CPU t-SNE in `manifoldsR` is already fast (Rust, SIMD kNN, FFT
optimiser on Unix). The GPU version's advantage is entirely in the kNN step,
so the payoff depends on how much of your wall-clock the kNN eats. Generally
speaking, if you have a larger data set, hit it with a GPU-accelerated kNN and
the FFT optimiser. Longer term, the idea is to also implement 2D-FFT that is
GPU-accelerated, but that is a larger project.

The kNN backend matters more here than in UMAP because the optimiser is
still CPU either way. `"nndescent"` and `"ivf"` are the workhorses,
`"exhaustive"` is for ground-truth checks or small data.

## Conclusions

Same mental model as CPU t-SNE, same knobs, same caveats (t-SNE is a
visualisation tool, cluster sizes and inter-cluster distances are not
meaningful, see the [manifoldsR vignette](https://gregorlueg.github.io/manifoldsR/articles/tsne.html)
for the details). Current pipeline is hybrid: kNN on GPU, optimiser on CPU.
Once the optimiser moves to GPU the whole thing runs on device end to end.
