---
title: "Leveraging GPU-acceleration in bixverse single cell"
vignette: >
  %\VignetteIndexEntry{Leveraging GPU-acceleration in bixverse single cell}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Gregor Lueg"
---

## Intro

This vignette walks through a standard single cell analysis on the PBMC3k data
set, demonstrating how `bixverse.gpu` can be used as a drop-in replacement for
the CPU-based neighbour search in `bixverse`. Everything up to and including PCA
is identical to the
[PBMC vignette](https://gregorlueg.github.io/bixverse/docs/articles/pbmc.html)
in the main package - the GPU-specific bits only come into play at the
neighbour search step. The core idea is to have GPU methods that are more
hardware agnostic, hence, the usage of 
[CubeCL with WGPU backend](https://github.com/tracel-ai/cubecl) to make the
code run on very different GPUs. Also, most of the time, it is quite overkill
and outside of massive data sets usually not worth it (or you always want
exhaustive, exact kNNs). If you have no idea what bixverse is and why some
of the code below looks quite unusual for single cell, please read 
[this](https://gregorlueg.github.io/bixverse/docs/articles/thinking_single_cell.html).

`bixverse.gpu` provides three GPU-accelerated kNN algorithms via a wgpu
backend:

- **Exhaustive**: Exact brute-force search. Precise, but scales quadratically.
  Best suited for smaller data sets or when you need exact neighbours.
- **IVF**: Inverted file index. Partitions the embedding space into Voronoi
  cells and probes only a subset at query time. Good throughput on larger data
  sets with a very minor precision trade-off.
- **CAGRA**: Builds a pruned navigational graph via NNDescent. Excellent
  recall on high-dimensional embeddings and very large data sets. Supports two
  retrieval modes: direct extraction from the NNDescent graph (faster, slightly
  less precise) or beam search over the pruned graph (slower, higher recall).

The exhaustive and IVF methods are exposed through `find_neighbours_gpu_sc()`,
whilst CAGRA has its own generic, `find_neighbours_cagra_sc()`, owing to its
distinct parameter set.

```{r}
#| label: setup
library(bixverse)
library(bixverse.gpu)
library(ggplot2)
library(data.table)
```

## Loading the data

Let's load the same data we know from `bixverse`.

```{r}
#| label: load pbmc3k

pbmc3k_path <- bixverse:::download_pbmc3k()

tempdir_pbmc <- tempdir()

sc_object <- SingleCells(dir_data = tempdir_pbmc)

mtx_io_params <- get_cell_ranger_params(pbmc3k_path)

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = mtx_io_params,
  streaming = FALSE,
  .verbose = TRUE
)

var <- get_sc_var(sc_object)

setnames_sc(
  object = sc_object,
  table = "var",
  old = "column1",
  new = "gene_symbol"
)

var <- get_sc_var(sc_object)

ensembl_to_symbol <- setNames(var$gene_symbol, var$gene_id)
symbol_to_ensembl <- setNames(var$gene_id, var$gene_symbol)
```

## Quality control

Same QC as per usual...

```{r}
#| label: gene set proportions

gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_symbol), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_symbol), gene_id]
)

sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  streaming = FALSE,
  .verbose = TRUE
)
```

Outlier detection:

```{r}
#| label: qc outlier detection

qc_df <- sc_object[[c("cell_id", "lib_size", "nnz", "MT")]]

metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)

directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)

qc <- run_cell_qc(metrics, directions, threshold = 3)
```

Set the labels to use.

```{r}
#| label: set cells to keep

sc_object[["outlier"]] <- qc$combined

cells_to_keep <- qc_df[!qc$combined, cell_id]

sc_object <- set_cells_to_keep(sc_object, cells_to_keep)
```

## Feature selection and PCA

Nothing GPU-specific here: HVG selection and PCA are identical to the CPU
pipeline. However, something important to see... The way the GPU code is 
written, the number of dimensions in the 

```{r}
#| label: hvg pca

sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 2000L,
  .verbose = TRUE
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 32L,
  sparse_svd = TRUE
)
```

## GPU-accelerated neighbour search

This is where `bixverse.gpu` comes in. Below we show all three methods. In
practice you would pick one - which one depends on the size of your data and
how much you care about exact neighbours versus speed.

### Exhaustive (exact, GPU)

For a data set as small as PBMC3k, exhaustive search on the GPU is the most
straightforward option.

```{r}
#| label: gpu exhaustive

sc_exhaustive <- find_neighbours_gpu_sc(
  object = sc_object,
  gpu_method = "exhaustive",
  k = 15L,
  dist_metric = "cosine",
  .verbose = TRUE
)
```

This might actually be slower than the CPU search on such a tiny data set due
to moving data from RAM to VRAM (if you are not on Apple Silicon) and GPU kernel
launch overhead. But it shows how it works... It becomes however much faster the
more data your GPU has to churn through.

### IVF (approximate, GPU)

IVF becomes worthwhile on larger data sets. The key tuning knobs are `nlist`
(number of Voronoi partitions) and `nprobe` (how many partitions to search at
query time). On PBMC3k the defaults are perfectly fine.

```{r}
#| label: gpu ivf

sc_ivf <- find_neighbours_gpu_sc(
  object = sc_object,
  gpu_method = "ivf",
  ivf_params = params_sc_ivf(k = 15L, ann_dist = "cosine"),
  .verbose = TRUE
)
```

### CAGRA (approximate, GPU)

CAGRA builds a navigational graph via NNDescent and then prunes it for
efficient beam search. Setting `extract_knn = TRUE` skips the beam search and
pulls the kNN directly from the NNDescent graph, which is faster but slightly
less precise.

```{r}
#| label: gpu cagra extract

sc_cagra_extract <- find_neighbours_cagra_sc(
  object = sc_object,
  cagra_params = params_sc_cagra(k_query = 15L, ann_dist = "cosine"),
  extract_knn = TRUE,
  .verbose = TRUE
)
```

With `extract_knn = FALSE`, the function runs beam search over the pruned CAGRA
graph. This is slower but yields higher recall, which can matter on larger,
higher-dimensional data.

```{r}
#| label: gpu cagra beam

sc_cagra_beam <- find_neighbours_cagra_sc(
  object = sc_object,
  cagra_params = params_sc_cagra(k_query = 15L, ann_dist = "cosine"),
  extract_knn = FALSE,
  .verbose = TRUE
)
```

From this point on, the `SingleCells` object carries the kNN and sNN graph
exactly as it would after a call to `find_neighbours_sc` (see bixverse 
function) - all downstream methods (clustering, UMAP, marker detection, 
etc.) work without modification.

## Conclusions

_"Will you add more GPU-accelerated methods?"_
Maybe... kNN searches were an obvious one as there are lots and lots of distance
calculations where GPUs have actually sufficient data to crunch. I will see
if other parts might make sense to push on the GPU: parametric UMAP could be
something to add here to train a model once and re-use repeatedly? TBD.

## Clean up

```{r}
#| label: clean up

unlink(tempdir_pbmc, recursive = TRUE, force = TRUE)
```
