---
title: "GPU-accelerated SCENIC"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{GPU-accelerated SCENIC}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Gregor Lueg"
---

## Intro

This vignette walks through the GPU-accelerated version of the SCENIC gene
regulatory network inference in `bixverse`. If you have not read the
[CPU SCENIC vignette](https://gregorlueg.github.io/bixverse/articles/bag_of_genes_single_cells.html)
first, do so. That vignette explains the biology and the CPU code path. This one 
focuses on what the GPU version buys you and how the API differs.

Behind `scenic_grn_sc_gpu()` sits a wave-scheduled, multi-tree GPU driver that 
dispatches ExtraTrees and RandomForest fits to the wgpu backend via
[cubecl](https://github.com/tracel-ai/cubecl). GRNBoost2 stays on the CPU 
because gradient boosting doesn't map cleanly onto the same multi-output
kernel design.

The `bixverse.gpu` package provides:

- **`scenic_grn_sc_gpu()`** for `SingleCells` and `MetaCells` objects. Streaming 
  and non-streaming drivers on the `SingleCells` path; a single
  in-memory driver on the `MetaCells` path.
- A single GPU-specific knob, `wave_byte_budget`, that caps VRAM use of the 
  per-wave histogram tensors. Default is 4 GiB. Shrink on 8 GB adapters shared 
  with other workloads; raise on 16 GB+ adapters to let the scheduler pick a 
  wider wave.

Everything else, the params, the target-gene shortlisting, the downstream motif 
enrichment, is imported straight from `bixverse`. The GPU version reuses the 
same `params_scenic()`, the same `scenic_gene_filter_sc()`, the same `ScenicGrn` 
result class, so downstream code (`identify_tf_to_genes`,
`tf_to_genes_correlations`, `tf_to_genes_motif_enrichment`) is identical.

```{r}
#| label: setup
library(bixverse)
library(bixverse.gpu)
library(data.table)
library(magrittr)
```

::: {.callout-note}
Vignettes were built locally on a MacBook Pro M1 Max. The GH runners were just
too slow and do not have proper GPU support. This gives an idea of speed on a 
decent, but older machine.
:::

## Rebuilding a processed PBMC3k object

We reconstruct the same `SingleCells` object used in the CPU SCENIC vignette. 
The chunk is folded because we care about SCENIC, not QC.

```{r}
#| label: rebuild-object
#| code-fold: true
#| code-summary: "Rebuild the PBMC3k object (click to expand)"
pbmc3k_path <- download_pbmc3k()
tempdir_pbmc <- file.path(tempdir(), "gpu_scenic")
dir.create(tempdir_pbmc, showWarnings = FALSE, recursive = TRUE)

sc_object <- SingleCells(dir_data = tempdir_pbmc)
mtx_io_params <- get_cell_ranger_params(pbmc3k_path)

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = mtx_io_params,
  mtx_streaming = FALSE,
  .verbose = FALSE
)

setnames_sc(sc_object, table = "var", old = "column1", new = "gene_symbol")

var <- get_sc_var(sc_object)
ensembl_to_symbol <- setNames(var$gene_symbol, var$gene_id)
symbol_to_ensembl <- setNames(var$gene_id, var$gene_symbol)

gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_symbol), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_symbol), gene_id]
)
sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  streaming = FALSE,
  .verbose = FALSE
)

qc_df <- sc_object[[c("cell_id", "lib_size", "nnz", "MT")]]
metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)
directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)
qc <- run_cell_qc(
  metrics = metrics,
  cells_to_keep = get_cells_to_keep(sc_object),
  directions = directions,
  threshold = 3
)
sc_object[["outlier"]] <- qc$combined
cells_to_keep <- qc_df[!qc$combined, cell_id]
sc_object <- set_cells_to_keep(sc_object, cells_to_keep)

sc_object <- find_hvg_sc(sc_object, hvg_no = 2000L, .verbose = FALSE)
sc_object <- calculate_pca_sc(sc_object, no_pcs = 30L, .verbose = FALSE)
sc_object <- find_neighbours_sc(
  sc_object,
  neighbours_params = params_sc_neighbours(knn = list(k = 15L)),
  .verbose = FALSE
)
```

## Gene filtering

Same as the CPU workflow. The gene filter is a cheap min-counts /
min-cells scan; there's no GPU version because there's no need for one.

```{r}
#| label: scenic-gene-filter
scenic_genes <- scenic_gene_filter_sc(
  object = sc_object,
  scenic_params = params_scenic(min_counts = 50L),
  .verbose = FALSE
)

length(scenic_genes)
```

## Transcription factor list

Aerts lab curated list, mapped to Ensembl.

```{r}
#| label: scenic-tfs
tf_dt <- data.table::fread(
  "https://resources.aertslab.org/cistarget/tf_lists/allTFs_hg38.txt",
  header = FALSE,
  col.names = "tf"
)
tf_dt[, gene_id := symbol_to_ensembl[tf]]
tf_dt <- tf_dt[!is.na(gene_id)]

nrow(tf_dt)
```

## GRN inference on the GPU

Same signature as `scenic_grn_sc()` from `bixverse`, with the extra
`wave_byte_budget` knob. Everything else, `scenic_params`, `tf_ids`,
`genes_to_take`, works the same way.

### ExtraTrees

`extratrees` is the recommended default on GPU. Random split thresholds mean 
each level dispatches a single split-scoring kernel, and the driver saturates 
the device with wide waves.

```{r}
#| label: gpu-scenic-et
scenic_res_et <- scenic_grn_sc_gpu(
  object = sc_object,
  tf_ids = tf_dt$gene_id,
  genes_to_take = scenic_genes,
  scenic_params = params_scenic(
    learner_type = "extratrees",
    gene_batch_size = 64L
  ),
  .verbose = TRUE
)

scenic_res_et
```

### RandomForest

`randomforest` uses exhaustive split evaluation. It's a bit slower per tree than 
ExtraTrees but often lands on tighter importance rankings.

```{r}
#| label: gpu-scenic-rf
scenic_res_rf <- scenic_grn_sc_gpu(
  object = sc_object,
  tf_ids = tf_dt$gene_id,
  genes_to_take = scenic_genes,
  scenic_params = params_scenic(
    learner_type = "randomforest",
    gene_batch_size = 64L
  ),
  .verbose = TRUE
)

scenic_res_rf
```

### GRNBoost2

Not available on GPU. The wrapper fails fast with a clear message before 
touching the device.

```{r}
#| label: gpu-scenic-grnboost2
#| error: true
scenic_grn_sc_gpu(
  object = sc_object,
  tf_ids = tf_dt$gene_id,
  genes_to_take = scenic_genes,
  scenic_params = params_scenic(learner_type = "grnboost2"),
  .verbose = FALSE
)
```

Fall back to `bixverse::scenic_grn_sc()` if you need it.

### Streaming vs non-streaming

On disk-backed `SingleCells` objects both drivers exist. Non-streaming loads the 
full sparse target column vector into host RAM up front and dispatches wave by 
wave. Streaming reads targets in I/O chunks and dispatches per chunk, capping 
peak host memory at roughly one chunk of sparse columns.

Set `streaming = FALSE` to force the non-streaming path when host RAM isn't the 
limiting factor:

```{r}
#| label: gpu-scenic-streaming
#| eval: false
scenic_res_streaming <- scenic_grn_sc_gpu(
  object = sc_object,
  tf_ids = tf_dt$gene_id,
  genes_to_take = scenic_genes,
  scenic_params = params_scenic(
    learner_type = "extratrees",
    gene_batch_size = 64L
  ),
  streaming = TRUE,
  .verbose = TRUE
)
```

Leave `streaming = NULL` to let the same `auto_streaming()` heuristic that the 
CPU version uses pick for you.

## TF-to-gene refinement

Because the result is a `ScenicGrn` object, the CPU downstream steps drop 
straight in.

```{r}
#| label: gpu-scenic-tf-to-gene
scenic_res_et <- identify_tf_to_genes(
  scenic_res_et,
  n_sd = 1,
  .verbose = TRUE
)

scenic_res_et <- tf_to_genes_correlations(
  x = scenic_res_et,
  object = sc_object,
  cor_filter = 0.01,
  .verbose = TRUE
)

tf_gene_dt <- get_tf_to_gene(scenic_res_et)
tf_gene_dt[, tf_symbol := ensembl_to_symbol[tf]]
tf_gene_dt[, gene_symbol := ensembl_to_symbol[gene]]

head(tf_gene_dt[order(-importance)], 5L)
```

## MetaCells path

The GPU driver also dispatches on `MetaCells`. Counts already live in memory as 
a sparse CSC block, so the streaming flag has no effect (and is silently 
ignored). Build the metacells the same way as the CPU vignette:

```{r}
#| label: gpu-scenic-metacells
mc_object <- generate_bt_meta_cells_sc(
  sc_object,
  sc_meta_cell_params = params_sc_bt_metacells(
    target_no_metacells = 200L
  ),
  .verbose = FALSE
)

scenic_res_mc <- scenic_grn_sc_gpu(
  object = mc_object,
  tf_ids = tf_dt$gene_id,
  scenic_params = params_scenic(
    learner_type = "extratrees",
    min_counts = 50L,
    min_cells = 0.02,
    gene_batch_size = 64L
  ),
  .verbose = TRUE
)

scenic_res_mc
```

Same result class, same downstream methods. `identify_tf_to_genes()`,
`tf_to_genes_correlations()` and `tf_to_genes_motif_enrichment()` all work
unchanged.

## CisTarget motif enrichment

Nothing changes on the CisTarget side. Same rankings, same annotations, same 
result. Not evaluated by default because the reference files are around 1.5 GB.

```{r}
#| label: gpu-scenic-cistarget
#| eval: false
paths <- download_cistarget_hg38()

rankings <- read_motif_ranking(paths$rankings)
annotations <- read_motif_annotation_file(paths$motif_annotations)

scenic_res_et <- tf_to_genes_motif_enrichment(
  x = scenic_res_et,
  motif_rankings = rankings,
  annot_data = annotations,
  gene_id_to_symbol = ensembl_to_symbol,
  cis_target_params = params_cistarget(nes_threshold = 3.0),
  only_high_conf_tf = FALSE,
  .verbose = TRUE
)

cistarget_dt <- get_cistarget_res(scenic_res_et)
head(cistarget_dt[order(-nes)])
```

## Choosing between CPU and GPU

Numbers first. M1 Max (8P+2E cores, 32 GPU cores, unified memory), 10k
cells, 1000 TFs, 4000 target genes in roughly 63 batches, 250 trees,
`max_depth = 10`, `min_samples_leaf = 50`:

| Learner      | CPU     | GPU    | Speed-up |
| ------------ | ------- | ------ | -------- |
| ExtraTrees   | 113.09s | 56.66s | 2.00x    |
| RandomForest | 87.02s  | 48.10s | 1.81x    |

That is end to end, gene filtering and batching included. Look at a
single 64-target batch against one CPU core and the GPU side is 21.3x
(ExtraTrees) and 23.6x (RandomForest). The end-to-end number is lower
because the CPU version parallelises across all cores, so you're really
comparing 32 GPU cores against 10 CPU cores. RandomForest ends up the
fastest absolute option here: `subsample_rate = 0.632` means each tree
only sees 63% of the cells.

Fidelity against the CPU implementation is per-target Pearson 0.987 for
RandomForest and 0.975 with bootstrapping on. Same trees, different
scheduling, so the importances agree to within sampling noise.

Where the GPU still loses:

- Small data (a few thousand cells, a few hundred targets). Fixed setup
  cost eats the win. Stay on CPU.
- GRNBoost2. Its single-target histogram-subtraction algorithm shares no
  structure across targets and gains nothing from a GPU port, so
  `scenic_grn_sc_gpu()` rejects it outright. Use
  `bixverse::scenic_grn_sc()`.
- RandomForest at very large cell counts. The per-wave tensors grow with
  `n_samples`, hitting roughly 8.4 GB at a million cells even at wave
  size 1, and the driver refuses rather than thrashing. ExtraTrees stays
  around 530 MB and is the path that scales. If you hit the ceiling,
  either switch learner or subsample cells via `n_subsample`.

`wave_byte_budget` is the knob for the middle ground. It caps VRAM for
the per-wave histogram and cumulative tensors. Shrink it on an 8 GB
adapter you're sharing with a display; raise it on 16 GB+ to let the
scheduler pick a wider wave.
