---
title: "Other GPU-accelerated methods"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Other GPU-accelerated methods}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

## Intro

Beyond the single cell stuff, `bixverse.gpu` has a small collection of
general-purpose GPU kernels that keep showing up: k-means, and Pearson
correlation / covariance on wide matrices. This vignette runs each one
against its CPU counterpart and gives a rough sense of when the GPU path
actually pays off.

Everything here runs on
[cubecl with the wgpu backend](https://github.com/tracel-ai/cubecl), so no
CUDA-only dependencies.

```{r}
#| label: setup
library(bixverse)
library(bixverse.gpu)
library(manifoldsR)
library(data.table)
library(ggplot2)
```

::: {.callout-note}
The correlation and covariance timing blocks below are set to `eval: false`
in the rendered docs. This vignette is built on GitHub Actions Linux
runners, which don't have real GPUs and fall back to software Vulkan via
Mesa (`WGPU_BACKEND=vulkan`). Any numbers produced there would be
misleading. Run the vignette locally on a machine with a proper GPU (or
Apple Silicon) to see the real comparison. This is also the reason why the 
timing eval is set to FALSE here.
:::

## GPU-accelerated k-means

k-means keeps coming back everywhere: quantisation, initialisation for other
clustering methods, IVF index construction, Harmony soft assignments. It's a
natural GPU workload because every iteration is dominated by a distance
kernel and a reduction.

`k_means_cluster_gpu()` is Lloyd's on the GPU. Knobs in
`params_kmeans_gpu()`:

- `k_means_init`: two variants exist on the GPU path, `"random"` (fast random
  selection) and `"parallel"` (also accepted as `"plusplus"`; this is
  k-means||, the parallel version of k-means++). `NULL` picks `"random"` when
  `n_centroids > 200`, `"parallel"` otherwise.
- `metric`: `"euclidean"` or `"cosine"`.
- `fixed`: run for a fixed number of iterations, skip the convergence check.
- `quantise`: hold the data buffer at fp16 on the GPU. Centroids and
  accumulators stay at fp32. Halves the data-buffer memory and boosts
  bandwidth on the assignment kernels. Needs `shader-f16` on the wgpu
  adapter.

10k points in 32D across 100 well-separated clusters:

```{r}
#| label: kmeans data
set.seed(42L)

cluster_data <- manifold_synthetic_data(
  type = "clusters",
  n_samples = 10000L,
  dim = 32L,
  parameters = params_clusters(n_clusters = 100L)
)
```

### CPU baseline

CPU reference is `manifoldsR::kmeans_cluster()`, which is Lloyd's along two
orthogonal acceleration axes: **Hamerly's bounds** (per-point upper/lower
distance bounds that prune most distance calculations each iteration) and a
**GEMM assignment path** (batch point-to-centroid distances through a `faer`
matmul when dim and centroid count are big enough). The four resulting
paths (`ParallelLloyd`, `GemmLloyd`, `HamerlySimd`, `HamerlyGemm`) are
auto-selected from the data shape. Hamerly falls back to plain Lloyd's
under Cosine (no triangle inequality).

Here we force both off. Comparison is against plain parallel SIMD Lloyd's,
same solution the accelerated paths converge to, just with a bigger
distance-calculation budget.

```{r}
#| label: kmeans cpu
km_cpu <- kmeans_cluster(
  data = cluster_data$data,
  k = 100L,
  method = "full",
  kmeans_params = params_kmeans(
    init = "random",
    metric = "euclidean",
    use_hamerly = FALSE,
    use_gemm = FALSE
  ),
  seed = 42L,
  .verbose = FALSE
)
```

### GPU version

Same interface, bump `k_means_iter` so both runs get a comparable budget:

```{r}
#| label: kmeans gpu
km_gpu <- k_means_cluster_gpu(
  data = cluster_data$data,
  k = 100L,
  kmeans_params = params_kmeans_gpu(
    k_means_iter = 100L,
    k_means_init = "random",
    metric = "euclidean",
    quantise = FALSE
  ),
  seed = 42L,
  .verbose = FALSE
)
```

k-means is a local optimiser sensitive to init, so the two won't be
bit-identical. Assignments should agree well though. ARI is the standard
measure:

```{r}
#| label: kmeans ari
ari_cpu_gpu <- calc_ari(km_cpu$assignments, km_gpu$assignments)
cat(sprintf("ARI (CPU vs GPU): %.3f\n", ari_cpu_gpu))
```

### Quantised GPU version

For memory-bound workloads (very wide feature matrices, very large N),
`quantise = TRUE` keeps the data buffer at `fp16` on the GPU while centroids
and accumulators stay at `fp32`. Frees up bandwidth on your GPU. Accuracy loss 
is (usually) small and easy to check:

```{r}
#| label: kmeans gpu quantised
km_gpu_q <- k_means_cluster_gpu(
  data = cluster_data$data,
  k = 100L,
  kmeans_params = params_kmeans_gpu(
    k_means_iter = 100L,
    k_means_init = "random",
    metric = "euclidean",
    quantise = TRUE
  ),
  seed = 42L,
  .verbose = FALSE
)

ari_gpu_q <- calc_ari(km_gpu$assignments, km_gpu_q$assignments)
cat(sprintf("ARI (GPU fp32 vs GPU fp16): %.3f\n", ari_gpu_q))
```

Tracks the full-precision run closely at a fraction of the memory.

### When to use the GPU version

On small data (a few thousand points, low dim) the CPU tends to win. Launch 
overhead and host-device transfers dominate. GPU pays off when:

- N is large (roughly > 50k),
- k is large (hundreds to thousands of centroids),
- Features are moderate to high dimensional.

## GPU-accelerated correlation and covariance

Pairwise Pearson correlation over a wide matrix looks trivial until you
notice it's a symmetric `X^T X` after centring and scaling. Textbook BLAS-3
GEMM. GPUs love this; modern CPU BLAS with SIMD also loves this, which is
why the crossover point isn't always where you'd expect.

`bixverse` already ships `rs_cor()`, which dispatches to
[`faer`](https://github.com/sarah-quinones/faer-rs) under the hood. faer is
a Rust BLAS with strong SIMD support and initial cubek-based GEMM only beat this
on very large data sets. However, `bixverse-rs` ships specific GPU kernels for 
correlations and co-variance calculations that actually beat `faer`. 
`bixverse.gpu` adds `rs_cor_gpu()` and `rs_cov_gpu()` for when you need have
your correlations fast on large data sets.

### Correlation

25k samples, 1k features. Base R vs CPU vs GPU:

```{r}
#| label: cor data
#| eval: false
set.seed(42L)

n_samples <- 50000L
n_features <- 2500L

random_data <- matrix(
  data = rnorm(n_samples * n_features),
  nrow = n_samples,
  ncol = n_features
)
```

Base R `cor()`:

```{r}
#| label: cor base
#| eval: false

# Base R will be slow...
t_base <- system.time({
  cor_base <- cor(random_data)
})
cat(sprintf("base R cor():   %.2fs\n", t_base[["elapsed"]]))
```

faer-backed CPU (via `bixverse`):

```{r}
#| label: cor cpu
#| eval: false
t_cpu <- system.time({
  cor_cpu <- rs_cor(random_data, spearman = FALSE)
})
cat(sprintf("rs_cor() CPU:   %.2fs\n", t_cpu[["elapsed"]]))
```

And GPU-accelerated:

```{r}
#| label: cor gpu
#| eval: false
t_gpu <- system.time({
  cor_gpu <- rs_cor_gpu(random_data, spearman = FALSE, verbose = FALSE)
})
cat(sprintf("rs_cor_gpu():   %.2fs\n", t_gpu[["elapsed"]]))
```

All three should agree:

```{r}
#| label: cor accuracy
#| eval: false
cat(sprintf("max |base - cpu|: %.2e\n", max(abs(cor_base - cor_cpu))))
cat(sprintf("max |cpu  - gpu|: %.2e\n", max(abs(cor_cpu - cor_gpu))))
```

CPU-vs-GPU deviation is fp32 GEMM rounding, not an algorithmic bug.

### Covariance

Same story:

```{r}
#| label: cov
#| eval: false
t_cpu_cov <- system.time({
  cov_cpu <- rs_covariance(random_data)
})
cat(sprintf("rs_covariance() CPU: %.2fs\n", t_cpu_cov[["elapsed"]]))

t_gpu_cov <- system.time({
  cov_gpu <- rs_cov_gpu(random_data, verbose = FALSE)
})
cat(sprintf("rs_cov_gpu():        %.2fs\n", t_gpu_cov[["elapsed"]]))

cat(sprintf("max |cpu - gpu|:     %.2e\n", max(abs(cov_cpu - cov_gpu))))
```

### When to use the GPU version

The GPU pulls ahead mainly on **very tall matrices**, lots of samples. Many
rows to sum over during centring and the GEMM, so the GPU's arithmetic
throughput outpaces what the CPU can sustain over host memory.

On Apple Silicon, faer's SIMD path is close enough that this crossover
sits fairly far to the right. Kernel launch and host-device transfer
overhead can wipe out the win on smaller inputs, and wgpu doesn't expose
proper tensor-core equivalents on that hardware, so you don't get the
blow-out margins you'd see on a CUDA machine.

Rough shape:

- Small feature count, modest sample size: `rs_cor()` on CPU wins.
- ~1k features, tens of thousands of samples: CPU and GPU close on Apple
  Silicon; discrete GPUs pull ahead earlier.
- Very tall or very wide: GPU wins clearly.

Same qualitative curve for `rs_cov_gpu()`.

## Conclusions

None of these are the headline GPU workloads (that's Harmony, PCA, kNN,
UMAP). They're the small pieces that make full pipelines feel snappy.
Correlation and covariance in particular sit inside a lot of downstream
methods (co-expression networks, PCA on covariance, contrastive PCA, ...),
so having a drop-in GPU version keeps the whole thing on one device.

More GPU utilities will land here over time. Watch this space.
